---
title: "generate output suitable for Phyloviz viz (Borbon Summer 2025)"
date: "2026-03-07"
output:
  html_document: default
  pdf_document: default
---


This Rmd parses mlst output
join with metadata file and produce result file: Result/meta_and_mlst.tsv
(for other analysis, figure/table generation)

create input files for Phyloviz:

Result/phylovizProfileDat_Borbon.tsv
Result/phylovizAuxDat_Borbon.tsv

These are 


TODO ++
====

There were 11 isolates that are not E.coli, 
search for "++" below and 
see DF mlstResultDF_unfiltered



~~~

Rmd
modeled after
- mission2022/guat_Gaby/code/ generate_dat4phyloviz.Rmd
- previous version "generate_dat4phyloviz_ee1352.Rmd" -->  create generated tree with cgST as node number, pie slides represented by source (bovine, feral pig, water, etc)

need to convert output from Mlst-Finder
to a format that Phyloviz expect.

also need metadata info, especially for Phyloviz Aux file.


mostly done, but need to discuss and decide places that have NA.  search for ++ below.


~~~


prep to use Phyloviz to create Min Spanning Tree (MST) using goeBURST-something algorithm
2026.0301

xref: usdastec/Env_EcO157/code/generate_dat4phyloviz_ee1352geo.Rmd




```{r setup}

library(pacman)
p_load(tidyverse)
p_load( dplyr )
p_load( readr )
p_load( stringr )


```


```{bash}

# path for MLST result
pwd
ls -la /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/metadata

```


# read inputs

```{r read_inputs}

# metadata info



metadataXlsx = readxl::read_xlsx( r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\metadata\GM_metadata.xlsx)", sheet="Sample IDs + Susceptibility", na="" )  # enclose windows path with r"(...)"  # strange syntax, but whatever.



metadataXlsx %>% dim()  

#metadataXlsx %>% View()
metadataXlsx %>% head(5)
metadataXlsx %>% colnames()


# read MLST result file
mlstResultFile       = "c:/tin/dataCache/Borbon_summer_25_Gabrielle/Result/mlst.all.tsv"
mlstResultTsv   = read_tsv( mlstResultFile, skip=0, show_col_types=F )        # sample records... isolate info    ## eventually to make to Aux file

mlstResultTsv %>% head(5)
```
```{r basic_sanity_check}

# any MLST that isn't e.coli?
mlstResultTsv %>% filter( PubMLST_scheme_name != "ecoli" )


# 11 of these.  discuss ++
# check phenotypic test, cross check that they are indeed not E.coli  (or predicted specie result that maybe avail in the annotation/ output by sequencer group)
# filter them out for now.

```

```{r massage_input_file}

# metadata, WgS_ID is saved as number, need it as string to join

metadataDF = metadataXlsx %>% 
  mutate( WGS_ID_as_string = as.character(WGS_ID) ) %>% 
  rename( Sample_Type_ID = Sample_Type ) %>% 
  rename( Isolate_Type = 'Sample Type' )


```

filenames eg: AI-1608_103_103_assembly.fasta
need to extract       ^^^
as join key to be able to merge metadata info.
use Regular Expression in str_match() to extract file id number

```{r prep_join}


mlstResultDF_unfiltered = mlstResultTsv %>% mutate(
  isolateKey = str_match( Filename, "AI[\\-]1608[\\_]([:digit:]+)[\\_]([:digit:]+)_assembly\\.fasta"  )[,2] 
)

mlstResultDF = mlstResultDF_unfiltered %>% 
  filter( PubMLST_scheme_name == "ecoli" )

mlstResultDF %>% dim()
mlstResultDF %>% select( Filename, isolateKey) %>% print(n=5)

```

```{r join_metadata_and_mlst}

meta_and_mlst_DF = full_join(
  x  = metadataDF,
  y  = mlstResultDF,
  by = c( `WGS_ID_as_string` = "isolateKey" )
)


meta_and_mlst_DF %>% head()
```
# write the combined mlst + metadata file into a new result file
```{r save_result}

write_tsv( meta_and_mlst_DF, "../Result/meta_and_mlst.tsv" )

```

# phyloviz seems to expect mlst file to have format like this:

 ST      gki     gtr     murI    mutS    recP    xpt     yqiZ
  1       10      6       6       6       12      13      8
  2       5       4       4       1       2       15      2
  3       5       3       4       1       6       2       1


so geneating a mlst_4_phyloviz.tsv  in the next block


```{r generate_phyloviz_profile_dat_file}

# ref /mission2022/guat_Gaby/code/generate_dat4phyloviz.Rmd

# mlst finder generated data from has these genes
#adk(6)  fumC(4) gyrB(32)  icd(1)  mdh(9)  purA(8) recA(6)
# regex escape: need to use \\( to match parenthesis
# note that another () need to be used to capture the number as [,2]
# mispatch likely result in NA in the cell.
# by def, MLST defined a ST number/id thing for it, the rest of the 7 column is just "showing work", they are always the same for a given ST
# phyloviz might not even need to look at them, this cleaning/conversion might not be necessary.

mlst_4_phyloviz_df = 
  meta_and_mlst_DF %>% 
  select( SeqType, Allele_ID_1, Allele_ID_2, Allele_ID_3, Allele_ID_4, Allele_ID_5, Allele_ID_6, Allele_ID_7 ) %>%
  unique() %>%
  mutate( #NEG = str_match( Allele_ID_1, "ctrl\\(([:digit:]+)\\)"  )[,2],   # negative control , expect no match
          ST   = SeqType,
          adk  = str_match( Allele_ID_1,  "adk\\(([:digit:]+)\\)"  )[,2],
          fumC = str_match( Allele_ID_2, "fumC\\(([:digit:]+)\\)"  )[,2],
          gyrB = str_match( Allele_ID_3, "gyrB\\(([:digit:]+)\\)"  )[,2],
          icd  = str_match( Allele_ID_4,  "icd\\(([:digit:]+)\\)"  )[,2],
          mdh  = str_match( Allele_ID_5,  "mdh\\(([:digit:]+)\\)"  )[,2],
          purA = str_match( Allele_ID_6, "purA\\(([:digit:]+)\\)"  )[,2],
          recA = str_match( Allele_ID_7, "recA\\(([:digit:]+)\\)"  )[,2]
          )


df4phylovizProfile = mlst_4_phyloviz_df %>%
  select( ST, adk, fumC, gyrB,  icd,  mdh,  purA, recA )

# check for NA, possible non e.coli gene list in mlst output  ++
df4phylovizProfile %>% 
  filter(if_any(everything(), is.na))    

# df4phylovizProfile %>% View()


df4phylovizProfile %>% dim() 

# above output imported into phyloviz, goeBURST full MLST generated a tree.... 
# (need to save result, see write_tsv below)



```
# phyloviz aux data would have the metadata 
pick only relevant column, too many columns may cause problem in phyloviz when data is loaded

```{r generate_phyloviz_aux_dat_file}

meta_and_mlst_DF %>% colnames()

df4phylovizAux = meta_and_mlst_DF %>% select( SeqType, `Isolate_Type`, House_ID )
df4phylovizAux %>% dim()

df4phylovizAux %>%
  filter(if_any(everything(), is.na)) 
# there are quite a number of NA, these are likely row that didn't end up matching as e.coli.
# ++ TBD what to do.
# maybe if they are indeed e.coli, force the scheme.
# 


# df4phylovizAux %>% View()


```


```{r tbd_save_phyloviz_dat_files}
write_tsv( df4phylovizProfile, "../Result/phylovizProfileDat_Borbon.tsv")
write_tsv( df4phylovizAux,     "../Result/phylovizAuxDat_Borbon.tsv")

```
