---
title: "generate output suitable for Phyloviz viz (USDA Env Samples EE1352) geo version"
date: "2026-02-14"
output:
  html_document: default
  pdf_document: default
---


Rmd
modeled after
- mission2022/guat_Gaby/code/ generate_dat4phyloviz.Rmd
- previous version "generate_dat4phyloviz_ee1352.Rmd" -->  create generated tree with cgST as node number, pie slides represented by source (bovine, feral pig, water, etc)

need to convert output from cgMlst(Finder?)
to a format that Phyloviz expect.
Hopefully in the ST format is okay, just more cols for cgMLST.

also need to filter input: EE1394 vs EE1352                           # done, use keepList
expect to be in 1352 isolates



~~~


prep to use Phyloviz to create Min Spanning Tree (MST) using goeBURST-something algorithm
2026.0301

xref: usdastec/Env_EcO157/code/generate_dat4phyloviz_ee1352geo.Rmd




```{r setup}

library(pacman)
p_load(tidyverse)
p_load( dplyr )
p_load( readr )
p_load( stringr )


```


```{bash}

# finding path for cgMLST result
# ls -la /mnt/c/tin/dataCache/stec_usda_Env_EcO157/Fasta4Prokka_EE1394/cgMlstFinder_Out_serial1_0823/Result/
pwd
#ls -la ../../../../dataCache/stec_usda_Env_EcO157/Fasta4Prokka_EE1394/cgMlstFinder_Out_serial1_0823/Result/

```


# read inputs

```{r read_inputs_1_metadata}

# metadata info

#EcO157_1394_filename = "../metadata/Environmental_EcO157_strains-Jay_G_UC_Berkeley.txt"  # converted from xlsx    # filename used in stInfo1394.Rmd, stored to Eco157_1394_df
isolateListFile       = "../metadata/Environmental_EcO157_strains-Jay_G_UC_Berkeley.txt"         # weasel, m01
isolateListTsv = read_tsv( isolateListFile )                 # sample records... isolate info    ## eventually to make to Aux file

isolateListTsv %>% dim()  # 1394 x 16
#isolate_keeplist %>% View()
isolateListTsv %>% head(5)
isolateListTsv %>% colnames()

```


```{r read_inputs_part_3_ST}

# the filter criteria from EE 1394 down to 1352 is based on ST 11
# code mostly from stInfo1394.Rmd

#ST_info_filename = "/mnt/c/tin/dataCache/stec_usda_Env_EcO157/Result_EE1394/MLST_OUT/mlst.all.tsv"
ST_info_filename = "../../../../dataCache/stec_usda_Env_EcO157/Result_EE1394/MLST_OUT/mlst.all.tsv"   # weasel
# header line needed manual fix to have tabs

ST_info_tsv = read_tsv( ST_info_filename, skip=0 )
ST_info_tsv %>% print( n=10 )


#xx geoXlsx = readxl::read_xlsx('C:\Users\tin61\Proton Drive\sn126\My files\Sn128\geo\Table 4 STEC O157 strains with location for Diana.xlsx', sheet="stains for tin to map", na="" )
geoXlsx = readxl::read_xlsx(r"(C:\Users\tin61\Proton Drive\sn126\My files\Sn128\geo\Table 4 STEC O157 strains with location for DianaV2.xlsx)", sheet="strains for tin to map", na="", guess_max=2000 )        # raw string, strange syntax // weasel use this
```
