---
title: "generate output suitable for Phyloviz viz (Borbon Summer 2025)"
date: "2026-03-01"
output:
  html_document: default
  pdf_document: default
---


Rmd
modeled after
- mission2022/guat_Gaby/code/ generate_dat4phyloviz.Rmd
- previous version "generate_dat4phyloviz_ee1352.Rmd" -->  create generated tree with cgST as node number, pie slides represented by source (bovine, feral pig, water, etc)

need to convert output from Mlst-Finder
to a format that Phyloviz expect.

also need metadata info, especially for Phyloviz Aux file.


~~~


prep to use Phyloviz to create Min Spanning Tree (MST) using goeBURST-something algorithm
2026.0301

xref: usdastec/Env_EcO157/code/generate_dat4phyloviz_ee1352geo.Rmd




```{r setup}

library(pacman)
p_load(tidyverse)
p_load( dplyr )
p_load( readr )
p_load( stringr )


```


```{bash}

# path for MLST result
pwd
ls -la /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/metadata

```


# read inputs

```{r read_inputs}

# metadata info



metadataXlsx = readxl::read_xlsx( r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\metadata\GM_metadata.xlsx)", sheet="Sample IDs + Susceptibility", na="" )  # enclose windows path with r"(...)"  # strange syntax, but whatever.



metadataXlsx %>% dim()  

#metadataXlsx %>% View()
metadataXlsx %>% head(5)
metadataXlsx %>% colnames()


# read MLST result file
mlstResultFile       = "c:/tin/dataCache/Borbon_summer_25_Gabrielle/Result/mlst.all.tsv"
mlstResultTsv   = read_tsv( mlstResultFile, skip=0, show_col_types=F )        # sample records... isolate info    ## eventually to make to Aux file

mlstResultTsv %>% head(5)
```
```{r basic_sanity_check}

# any MLST that isn't e.coli?
mlstResultTsv %>% filter( PubMLST_scheme_name != "ecoli" )


# 11 of these.  discuss ++
# filter them out for now.

```

```{r massage_input_file}

# metadata, WgS_ID is saved as number, need it as string to join

metadataDF = metadataXlsx %>% mutate( 
  WGS_ID_as_string = as.character(WGS_ID) 
  )


```

filenames eg: AI-1608_103_103_assembly.fasta
need to extract       ^^^
as join key to be able to merge metadata info.
use Regular Expression in str_match() to extract file id number

```{r prep_join}


mlstResultDF_unfiltered = mlstResultTsv %>% mutate(
  isolateKey = str_match( Filename, "AI[\\-]1608[\\_]([:digit:]+)[\\_]([:digit:]+)_assembly\\.fasta"  )[,2] 
)

mlstResultDF = mlstResultDF_unfiltered %>% 
  filter( PubMLST_scheme_name == "ecoli" )

mlstResultDF %>% dim()
mlstResultDF %>% select( Filename, isolateKey) %>% print(n=5)

```

```{r join_metadata_and_mlst}

meta_and_mlst_DF = full_join(
  x  = metadataDF,
  y  = mlstResultDF,
  by = c( `WGS_ID_as_string` = "isolateKey" )
)


meta_and_mlst_DF %>% head()
```
# write the combined mlst + metadata file into a new result file
```{r save_result}

write_tsv( meta_and_mlst_DF, "../Result/meta_and_mlst.tsv" )

```

# phyloviz seems to expect mlst file to have format like this:

 ST      gki     gtr     murI    mutS    recP    xpt     yqiZ
  1       10      6       6       6       12      13      8
  2       5       4       4       1       2       15      2
  3       5       3       4       1       6       2       1


so geneating a mlst_4_phyloviz.tsv  in the next block
```{r generate_phyloviz_profile_dat_file}

df4phylovizProfile = meta_and_mlst_DF %>% select(
  SeqType, Allele_ID_1, Allele_ID_2, Allele_ID_3, Allele_ID_4, Allele_ID_5, Allele_ID_6, Allele_ID_7 
) %>% unique() %>% drop_na()

```

```{r generate_phyloviz_aux_dat_file}
```


```{r tbd_save_phyloviz_profile_dat_file}
write_tsv( df4phylovizProfile, "../Result/phylovizProfileDat_Borbon.tsv")

```
