---
title: "generate output suitable for Phyloviz viz (Borbon Summer 2025)"
date: "2026-04-11"
output:
  html_document: default
  pdf_document: default
---


create input files for grapeTree to create Min Spanning Tree


This end up being pretty different than previous meta+mlst+phyloviz.Rmd
workflow changed in significant details.
now:

- specie determined by rMLST
- MLST re-ran with forced e.coli to get ST, (but isolate filtered for non-EC from rMLST result)
- use meta_and_mlst and split into CFX positive vs negative (selective media)
- create grapetree input files for CFX+ vs CFX-


(
ran rMLST to determine organism, non E.coli will be fitlered out.  result of rMLST placed in new version of metadata file
GM all samples + species 
https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
[saved to metadata/GM_all_samples+species.tsv]
)

create input files for grapeTree:

Result/grapeTree_ProfileDat_Borbon_CTXyes.tsv
Result/grapeTree_AuxDat_Borbon_CTXyes.tsv

Result/grapeTree_ProfileDat_Borbon_CTXno.tsv
Result/grapeTree_AuxDat_Borbon_CTXno.tsv


PS.  Unsure what is abreviation of Ceftriaxone, could be CFX, CRO, CTX.

but this article says CFX, and it is most clear mnemonic to me, so using that in this code.
but for paper, double check!

https://pmc.ncbi.nlm.nih.gov/articles/PMC5968784/#:~:text=Ceftriaxone%20(CFX)%20and%20clavulanic%20acid%20(CA)%20are,properties%20have%20been%20published%20in%20recent%20years.



~~~

Rmd
modeled after

- usdastec/Env_EcO157/code/generate_dat4phyloviz_ee1352geo.Rmd
- xxx mission2022/guat_Gaby/code/ generate_dat4phyloviz.Rmd
- xxx previous version "generate_dat4phyloviz_ee1352.Rmd" -->  create generated tree with cgST as node number, pie slides represented by source (bovine, feral pig, water, etc)

need to convert output from Mlst-Finder
to a format that Phyloviz expect.

also need metadata info, especially for Phyloviz Aux file.



~~~


prep to use GrapeTree to create Min Spanning Tree (MST) 




```{r setup}

library(pacman)
p_load(tidyverse)
p_load( dplyr )
p_load( readr )
p_load( stringr )


```


```{bash}

# path for MLST result
pwd
ls -la /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/metadata

```


# read inputs

```{r read_inputs}


# metadata info
# still using this metadata file, need to extract CFX+ vs CFX- from "Sample Type" col
metadataXlsx = readxl::read_xlsx( r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\metadata\GM_metadata.xlsx)", sheet="Sample IDs + Susceptibility", na="" )  # enclose windows path with r"(...)"  # strange syntax, but whatever.
metadataXlsx %>% dim()  
#metadataXlsx %>% View()
#metadataXlsx %>% head(5)
metadataXlsx %>% colnames()



# read MLST result file (now use the forced E.coli method)
#                   /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv
mlstResultFile       = "c:/tin/dataCache/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv"   
mlstResultTsv   = read_tsv( mlstResultFile, skip=0, show_col_types=F )        # sample records... isolate info    ## eventually to make to Aux file

mlstResultTsv %>% head(5) 
mlstResultTsv %>% colnames()

# specie info from rMLST 
# from MLST, and some are not e.coli.  determination of e.coli will now be determined by rMLST, 
# with info manually merged into the metadata file 
# GM all samples + species 
# https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
# mostly treat this file for rMLST specie determination using column rMLST_result_full_run
# the `MLST results` was determined when run_mlst.sh let the program determine specie
# (which will be changed by this incarnation/version of this script)
metadataWspecieTsv = read_tsv( "../metadata/GM_all_samples+species.tsv" )
metadataWspecieTsv %>% dim()  
metadataWspecieTsv %>% colnames()  

```



```{note eval=F}

# run rMLST on fasta that MLST said is not ecoli

# cmd from
#  cat nonEcoliPerMlst.tsv | awk -F. '{print "/global/home/users/tin/gs/opt/rmlst/rmlst.py --file " $1 ".fasta | tee " $1 ".rMlst.txt " }' > run_rmlst.sh
/global/home/users/tin/gs/opt/rmlst/rmlst.py --file AI-1608_136_136_assembly.fasta | tee AI-1608_136_136_assembly.rMlst.txt



# rmlst.py is from https://pubmlst.org/species-id/species-identification-via-api#python

# run on n0062.s4


only iso #136 was sure as E.coli, 
iso #203 was only at Genus level as Escherichia 

[tin@n0062 Borbon_summer_25_Gabrielle]$ cat AI-1608_136_136_assembly.rMlst.txt
Rank: SPECIES
Taxon:Escherichia coli
Support:100%
TaxonomyPseudomonadota > Gammaproteobacteria > Enterobacterales > Enterobacteriaceae > Escherichia > Escherichia coli


[tin@n0062 Borbon_summer_25_Gabrielle]$ cat AI-1608_203_203_assembly.rMlst.txt
Rank: GENUS
Taxon:Escherichia
Support:100%
TaxonomyPseudomonadota > Gammaproteobacteria > Enterobacterales > Enterobacteriaceae > Escherichia


or see  run_rmlst.TEE_OUT   for screen output of all rMlst.py checked fasta seq.

from mlst.all.tsv   /global/home/users/tin/gs/dataCache/_Doug_Gabrielle_WGS_Data/Summer_2025_WGS_Sequences/AI-1608_results/AI-1608_assembly/Borbon_summer_25_Gabrielle/MLST_OUT

AI-1608_136_136_assembly.fasta  cfreundii       -       aspC(171)       clpX(193)       fadD(112)       mdh(172)        arcA(~114)      dnaG(84)        lysP(101)
AI-1608_203_203_assembly.fasta  -       -

  
  full_updated_summary-AI-1608.csv  # that came with zip file with the sequences:  
  
Sample	Name	user_sample_name	Best match	Similarity	ANI	total_contigs	largest_contig_length_Mb	largest_contig_coverage	Total Mbp Sequenced	Completeness	Contamination
34	136	136	GCF_003697165.2 Escherichia coli DSM 30083 = JCM 1649 = ATCC 11775	28.12%	97.38%	1	4.621	46	1343.42	100	0.06
101	203	203	GCF_003697165.2 Escherichia coli DSM 30083 = JCM 1649 = ATCC 11775	17.85%	96.22%	632	0.025	5	39.2	50.2	2.13



```

```{r massage_input_file, eval=T}

# metadata, WgS_ID is saved as number, need it as string to join
metadataDF = metadataXlsx %>% 
  mutate( WGS_ID_as_string = as.character(WGS_ID) ) %>% 
  rename( Sample_Type_ID = Sample_Type ) %>% 
  rename( Isolate_Type = 'Sample Type' )


# GM all samples + species 
# https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
# mostly treat this file for rMLST specie determination using column rMLST_result_full_run
# the `MLST results` was determined when run_mlst.sh let the program determine specie
# (which will be changed by this incarnation/version of this script)
rMLST_specie = metadataWspecieTsv %>% 
  rename( rMLST_specie = rMLST_result_full_run ) %>%
  select( - `rMLST results` )  # this col was manually created G and has some inconsistency

```

```{r rMLST_filter_nonEC}

# rMLST_specie %>% View()
rMLST_specie %>% dim()   # 104 rows
rMLST_EC    = rMLST_specie %>% filter( rMLST_specie == "Escherichia coli" )

rMLST_EC %>% dim() #  93 rows ie filtered 11 rows (incl 1 NA)
rMLST_EC %>% colnames()
# "Name" col has the isolate ID that match Gabrielle's "user_sample_name" (these 2 cols seems to match, except 126 missing in user_sample_name)

rMLST_nonEC <- rMLST_specie %>% filter( ! grepl( "Escherichia coli", rMLST_specie   ) )  
rMLST_nonEC %>% dim() #  expect 11 !!   (1 is NA)
rMLST_nonEC %>% print( n=12)


```



filenames eg: AI-1608_103_103_assembly.fasta
need to extract       ^^^
as join key to be able to merge metadata info.
use Regular Expression in str_match() to extract file id number


mlst output examples

mlst   AI-1608_206_206_assembly.fasta
AI-1608_206_206_assembly.fasta  ecoli   10      adk(10) fumC(11)        gyrB(4) icd(8)  mdh(8)  purA(8) recA(2)

mlst  --scheme ecoli AI-1608_206_206_assembly.fasta
AI-1608_206_206_assembly.fasta  ecoli   10      adk(10) fumC(11)        gyrB(4) icd(8)  mdh(8)  purA(8) recA(2)


mlst --legacy --scheme ecoli AI-1608_206_206_assembly.fasta
FILE    SCHEME  ST      adk     fumC    gyrB    icd     mdh     purA    recA
AI-1608_206_206_assembly.fasta  ecoli   10      10      11      4       8       8       8       2



--legacy is easiest to parse
grapetree and phyloviz likely just want the number (for matrix)

/mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv


```{r prep_joinKeys}

mlstResultDF_unfiltered = mlstResultTsv %>% mutate(
  isolateKey = str_match( FILE, "AI[\\-]1608[\\_]([:digit:]+)[\\_]([:digit:]+)_assembly\\.fasta"  )[,2] 
) %>%  rename( forced_EC_scheme = SCHEME )   # to be very clear MLST output were all forced to E.coli

```


```{r join_metadata_and_mlst}

meta_and_mlst_DF = full_join(
  x  = metadataDF,
  y  = mlstResultDF_unfiltered, #mlstResultDF,
  by = c( `WGS_ID_as_string` = "isolateKey" )
)


meta_and_mlst_DF %>% head()
meta_and_mlst_DF %>% dim()   # 104... some might not be EC
#meta_and_mlst_DF %>% View()

```

```{r join_rMlst}

# filtered rMLST, right join, so non EC would be removed
# expect 93 rows

meta_and_EC_ST = right_join(
  x  = meta_and_mlst_DF,
  y  = rMLST_EC,
  by = c( "WGS_ID" = "Name")
)

meta_and_EC_ST %>% dim()  # 93 rows, good
meta_and_EC_ST %>% colnames() 
```

# write the combined mlst + metadata file into a new result file
```{r save_result}

##old previos version## write_tsv( meta_and_mlst_DF, "../Result/meta_and_mlst.tsv" )


## saving new "v2" result // this is optional, in case she wants to look at it.  __key_DF__ __DW_DF__

# write_tsv( meta_and_EC_ST, "../Result/meta_and_EC_ST.tsv" )  
# avoid saving result to public git repo till paper is published
# so save to temporary location and upload to gDrv: https://drive.google.com/drive/u/0/folders/1v-GvAj3N6EZ4LbPpRg1X9DUBhLPn-wFw
write_tsv( meta_and_EC_ST,  r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\Result\joinedRmdDF.tsv)" ) 

```

## split CFX+ vs CFX-  


```{r split_cfx_pos_vs_neg}


meta_and_EC_ST__CfxPos = meta_and_EC_ST %>%
  filter(   grepl( "Ceftriaxone", Isolate_Type, ignore.case=T ) ) 

meta_and_EC_ST__CfxNeg = meta_and_EC_ST %>%
  filter( ! grepl( "Ceftriaxone", Isolate_Type, ignore.case=T ) ) 


meta_and_EC_ST__CfxPos %>% dim()  # 37
meta_and_EC_ST__CfxNeg %>% dim()  # 56

# meta_and_EC_ST__CfxNeg %>% View()
```

```{r prepGrapeTreeDf}

# Ceftriaxone Positive

grapeTree_1prof_Borbon_CfxPos = meta_and_EC_ST__CfxPos %>% select( ST, adk,	fumC,	gyrB,	icd,	mdh,	purA,	rec ) %>%
  rename(`#Name` = "ST" ) 
  
# metadata
grapeTree_2aux_Borbon_CfxPos = meta_and_EC_ST__CfxPos %>% select( ST, Isolate_Type, rMLST_specie ) %>%
  rename( `id` = "ST")

# Ceftriaxone Negative

grapeTree_1prof_Borbon_CfxNeg = meta_and_EC_ST__CfxNeg %>% select( ST, adk,	fumC,	gyrB,	icd,	mdh,	purA,	rec ) %>%
  rename(`#Name` = "ST" ) 
  
# metadata
grapeTree_2aux_Borbon_CfxNeg = meta_and_EC_ST__CfxNeg %>% select( ST, Isolate_Type, rMLST_specie ) %>%
  rename( `id` = "ST")


```

# actually save file for grapetree

```{r write_result_file_in_grapeTree_format, eval=F}
# save only when sure

write_tsv( grapeTree_1prof_Borbon_CfxPos, "../RESULT_UPLOAD/grapeTree_1prof_Borbon_CfxPos.tsv" )
write_tsv( grapeTree_2aux_Borbon_CfxPos,  "../RESULT_UPLOAD/grapeTree_2aux_Borbon_CfxPos.tsv")

write_tsv( grapeTree_1prof_Borbon_CfxNeg, "../RESULT_UPLOAD/grapeTree_1prof_Borbon_CfxNeg.tsv" )
write_tsv( grapeTree_2aux_Borbon_CfxNeg,  "../RESULT_UPLOAD/grapeTree_2aux_Borbon_CfxNeg.tsv")

```


# some sanity check
why all node are the same size in grapetree?
see ST frequency distribution 
```{r ST_freq}

meta_and_EC_ST__CfxPos %>% group_by( `ST` ) %>% summarize( ST_count = n())  %>% arrange( desc(ST_count) )
meta_and_EC_ST__CfxNeg %>% group_by( `ST` ) %>% summarize( ST_count = n())  %>% arrange( desc(ST_count) )

```

### below for phyloviz, remove when done with grapetree



# phyloviz seems to expect mlst file to have format like this:
# (mlst --legacy would have number only for the matrix without the gene name repeated)
# or see meta+mlst+phyloviz.Rmd for R code that strip out the characters 

 ST      gki     gtr     murI    mutS    recP    xpt     yqiZ
  1       10      6       6       6       12      13      8
  2       5       4       4       1       2       15      2
  3       5       3       4       1       6       2       1


so geneating a mlst_4_phyloviz.tsv  in the next block


```{r generate_phyloviz_profile_dat_file}
# meta+mlst+phyloviz.Rmd  has some code
# but a bit diff now with this new workflow
# write new code if needed.
```
