---
title: "generate Pathotype table (Borbon Summer 2025)"
date: "2026-06-06 2026-10-03"
output:
  html_document: default
  pdf_document: default
---

version

2026-06ish: MLST and/or Abricate, simple result for GM.
2026-07ish: pathotype table (but never finished here, see irrExp26 instead)
2026-1003:  heatmap for figures Jay want for short paper (PLOS?)


~~~

(code based on meta+mlst+grapetree.Rmd)

Generate pathotype table for Gabrielle's Borbon Summer 2025 dataset
(Jay suggestion bmail 2026.06~05)
Never completed that here, but went to work for irrExp26, so see code there.

dataCache: 
/mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT   // weasel
~tin/dataCache/... // m02


use Abricate_vfdb_summary.csv
don't use mlst.all.tsv ?   but the xlsx used below (for grapetree generation also).

eg desired result table:

Pathotype\source  mom  |  child   |  water
---------  ------------|----------|---------
ExPEC           n (%)  |  n (%)   |  n (%)
ETEC            n (%)  |  n (%)   |  n (%)
EPEC            n (%)  |  n (%)   |  n (%)
EAEC            n (%)  |  n (%)   |  n (%)
DAEC            n (%)  |  n (%)   |  n (%)
EIEC/Shigella   n (%)  |  n (%)   |  n (%)  
STEC/EHEC       n (%)  |  n (%)   |  n (%)      # think STEC would be 0



~~~

Does it matter if Ceftriaxone +ve or -ve (CRO+ vs CRO-)? 
Separate table for the two media?
A> Combined for manuscript, separated for self evaluation and highlight interesting points in discussion.  maybe as supplements?



Gabrielle, whatever write up that you already have, if you can send me a copy, that would help me get more context and remember where we left off :D 


++ Need to check denominator count.  E.coli?  what rows had been filtered?  ++

inputs:
- meta data for sample source, CRO+ vs CRO- (from "Sample Type" col in the xlsx file, but need to extract CRO+ vs CRO- info from the string),
- abricate vfg (not resfinder) result   for the genes used to define Pathotype.
  https://drive.google.com/file/d/1DD_T15Uz0C_zydz8EeNN8fbX_MPqG5__/view?usp=drive_link 
  /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv    104 rows

~~~~

ExPEC definition.  Use local def from Liseth.  citation?

other Pathotype def: 
Paper by Kelsey J. Jesser, Karen Levy: 
Curr Opin Infect Dis. 2020 October ; 33(5): 372–380. doi:10.1097/QCO.0000000000000665

My annotated copy:
J:\Shared drives\Tin_Ho_gDrv_BL_shareDrv\mph_hw\mph_prj_bioinfo\ExPEC_Ala\ Defining diarrheagenic E. coli.pdf 
https://drive.google.com/drive/u/0/folders/1nKbMtrcIxkGjHMCUsdoQvPtEL1MiLbPS


~~~

gDrive for Borbon_summer_25:
https://drive.google.com/drive/u/0/folders/1eDXqacdExuWbvYvnBdqHIjb1QMkfTAJZ

/global/home/users/tin/gs/dataCache/_Doug_Gabrielle_WGS_Data/Summer_2025_WGS_Sequences/AI-1608_results/AI-1608_assembly/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC

(
previously 
ran rMLST to determine organism, non E.coli will be fitlered out.  result of rMLST placed in new version of metadata file
GM all samples + species 
https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
[saved to metadata/GM_all_samples+species.tsv]
)



~~~



```{r setup}

library(pacman)
p_load(tidyverse)
p_load( dplyr )
p_load( readr )
p_load( stringr )


```


```{bash}

# path for MLST result
pwd
ls -la /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/metadata

ls -la /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv    
# 104 rows

```


# Fetch files, esp for m02, as sshfs require installing system extension that require reboot...
(weasel alread had downloaded it)

cd /Users/tin/dataCache/Borbon_summer_25_Gabrielle/Result
scp -p dtn.brc.berkeley.edu:/global/home/users/tin/gs/dataCache/_Doug_Gabrielle_WGS_Data/Summer_2025_WGS_Sequences/AI-1608_results/AI-1608_assembly/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv .

```{r define_filenames_path}

#GM_metadata_path =  r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\metadata\GM_metadata.xlsx)"  # weasel
GM_metadata_path =  "~/dataCache/Borbon_summer_25_Gabrielle/metadata/GM_metadata.xlsx"          # m02 bigmac02
# /Users/tin/dataCache/Borbon_summer_25_Gabrielle/metadata


# read MLST result file (now use the forced E.coli method)
#                   /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv
#mlstResultFile       = "c:/tin/dataCache/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv"
#mlstResultFile       = "c:/tin/dataCache/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv"
mlstResultFile   = "~/dataCache/Borbon_summer_25_Gabrielle/Result/mlst.all.forceEC.tsv"   # m02, fetched from dtn 2026.1003, file dataed 4/11
## the .raw. version had the right header that isn't duplicated string, forceEC added header, didnt have to fix header in my script.

# specie info from rMLST 
metadataWspeciePath = "../metadata/GM_all_samples+species.tsv"

#VfdbResultFile = "/mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv"
#VfdbResultFile = "c:/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv"     # wide version, weasel
VfdbResultFile     = "~tin/dataCache/Borbon_summer_25_Gabrielle/Result/Abricate_vfdb_summary.csv"          # wide version, m02
VfdbResultFile_lng = "~tin/dataCache/Borbon_summer_25_Gabrielle/Result/Abricate_vfdb_combined.csv"         # wide version, m02

resfResultFile     = "~tin/dataCache/Borbon_summer_25_Gabrielle/Result/Abricate_resfinder_summary.csv"     # wide version, m02
resfResultFile_lng = "~tin/dataCache/Borbon_summer_25_Gabrielle/Result/Abricate_resfinder_combined.csv"    # long version, m02

# gDrive share https://drive.google.com/drive/folders/1eDXqacdExuWbvYvnBdqHIjb1QMkfTAJZ
# cd /global/home/users/tin/gs/dataCache/_Doug_Gabrielle_WGS_Data/Summer_2025_WGS_Sequences/AI-1608_results/AI-1608_assembly/Borbon_summer_25_Gabrielle


```
# read inputs

```{r read_inputs}

# metadata info
# still using this metadata file, need to extract CFX+ vs CFX- from "Sample Type" col
#metadataXlsx = readxl::read_xlsx( r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\metadata\GM_metadata.xlsx)", sheet="Sample IDs + Susceptibility", na="", col_types="text" )  # HORRIBLE! col_types="text"  convert WGS_ID to 103.0 !!  wtf?!
#metadataXlsx = readxl::read_xlsx( r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\metadata\GM_metadata.xlsx)", sheet="Sample IDs + Susceptibility", na="" )  # enclose windows path with r"(...)"  # strange syntax, but whatever.
metadataXlsx = readxl::read_xlsx( GM_metadata_path, sheet="Sample IDs + Susceptibility", na="" )  # enclose windows path with r"(...)"  # strange syntax, but whatever.


metadataXlsx %>% dim()  
#metadataXlsx %>% View()
metadataXlsx %>% head(5)
metadataXlsx %>% colnames()
metadataXlsx %>% str()


mlstResultTsv   = read_tsv( mlstResultFile, skip=0, show_col_types=F, col_types=cols(.default="c") )        # sample records... isolate info    ## eventually to make to Aux file
# col_types=cols(.default="c") force to character, no need to parse any of the col as number

#mlstResultTsv |> str()

# specie info from rMLST 
# from MLST, and some are not e.coli.  determination of e.coli will now be determined by rMLST, 
# with info manually merged into the metadata file 
# GM all samples + species 
# https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
# mostly treat this file for rMLST specie determination using column rMLST_result_full_run
# the `MLST results` was determined when run_mlst.sh let the program determine specie
# (which will be changed by this incarnation/version of this script)
# metadataWspecieTsv = read_tsv( "../metadata/GM_all_samples+species.tsv", skip=0, show_col_types=F, col_types=cols(.default="c") )  # here type forcing is fine, still have 103 without .0 added. phew!

metadataWspecieTsv = read_tsv( metadataWspeciePath, skip=0, show_col_types=F, col_types=cols(.default="c") )  # here type forcing is fine, still have 103 without .0 added. phew!


metadataWspecieTsv %>% colnames()  
metadataWspecieTsv %>% dim()       # 104 x 9
metadataWspecieTsv |>  str()
metadataWspecieTsv |>  head()

VfdbResultTsv  = read_tsv( VfdbResultFile, skip=0, show_col_types=F, col_types=cols(.default="c") )           # NUM_FOUND could be a number, but not really computing on it anyway

VfdbResultTsv |> head(5) 
#VfdbResultTsv |> colnames()
VfdbResultTsv |> dim()      # 103 rows, 201 cols


VfdbResultTsvLng  = read_tsv( VfdbResultFile_lng, skip=0 ) #, show_col_types=F, col_types=cols(.default="c") )           # NUM_FOUND not in long version

VfdbResultTsvLng |> filter( str_detect( `%COVERAGE`, regex(";")  ) )  # no match, so bonafide just a number.  no need to force col_types
VfdbResultTsvLng |> filter( str_detect( `%IDENTITY`, regex(";")  ) )  # ditto
sum( VfdbResultTsvLng$`%COVERAGE`)  # can sum it, no error, likely no funny multi-char biz.  
sum( VfdbResultTsvLng$`%IDENTITY`)  # can sum it, no error, likely no funny multi-char biz.  



resfResultTsv = read_tsv( resfResultFile, skip=0, show_col_types=F, col_types=(.default="c") )
#resfResultTsv |> str()  # 103*66
# resfResultTsv |> View()  

# the combined.tsv file is long version of summary.tsv, and has more info (cols)
# GENE is what used to be colnames in the wide(summary) version.
# %IDENTITY 
# %COVERAGE

resfResultTsvLng = read_tsv( resfResultFile_lng, skip=0 ) #, show_col_types=F, col_types=(.default="c") )
resfResultTsvLng |> str()  # 497*15
# resfResultTsvLng |> View()  

resfResultTsvLng |> filter( str_detect( `%COVERAGE`, regex(";|/|\ ")  ) )  # no match, so bonafide just a number.  no need to force col_types
resfResultTsvLng |> filter( str_detect( `%IDENTITY`, regex(";")  ) )  # ditto
resfResultTsvLng |> filter( is.na( `%COVERAGE`) )   # no NA.  good.  
sum( resfResultTsvLng$`%COVERAGE`)  # can sum it, no error, likely no funny multi-char biz.  
sum( resfResultTsvLng$`%IDENTITY`)  # can sum it, no error, likely no funny multi-char biz.  

```



# Abricate output: the combined.tsv file is long version of summary.tsv, and has more info (cols)
GENE is what used to be colnames in the wide(summary) version.

these are additional col in the long version that isn't in the wide(summary) version:
PRODUCT	aminoglycoside O-phosphotransferase APH(3')-IIIa	Gene product (if available)
RESISTANCE	TETRACYCLINE;FUSIDIC_ACID	putative antibiotic resistance phenotype, ;-separated


%COVERAGE	100.00%	Proportion of gene covered   ## this number is what goes to summary.tsv result
%IDENTITY	99.95%	Proportion of exact nucleotide matches in the covered part


ABRicate can combine results into a simple matrix of gene presence/absence. 
An absent gene is denoted . and a present gene is represented by its '%COVERAGE`. 
This can be individual abricate reports, or a combined one.

https://github.com/tseemann/ABRICATE


```{note eval=F}

# run rMLST on fasta that MLST said is not ecoli

```

```{r massage_input_file, eval=T}

# metadata, WgS_ID is saved as number, need it as string to join
metadataDF = metadataXlsx %>% 
  mutate( WGS_ID_as_string = as.character(WGS_ID) ) %>%   # forced input col type didn't work, so keep this.
  rename( Sample_Type_ID = Sample_Type ) %>% 
  rename( Isolate_Type = 'Sample Type' ) |>
  mutate(
    source_water  = if_else( grepl( "Water",  Isolate_Type, ignore.case=T),  1, 0, missing=0 ),
    source_mother = if_else( grepl( "Mother", Isolate_Type, ignore.case=T),  1, 0, missing=0 ),
    source_child  = if_else( grepl( "Child",  Isolate_Type, ignore.case=T),  1, 0, missing=0 )
  ) |> mutate(
    QA_source_count = ( source_water + source_mother + source_child )  # expect 1 for every row, check if not!
  )


metadataDF |> filter( QA_source_count != 1 )   # they are indeed all 1, pattern match worked.
#metadataDF |> head(5)
#metadataDF |> View()



# GM all samples + species 
# https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
# mostly treat this file for rMLST specie determination using column rMLST_result_full_run
# the `MLST results` was determined when run_mlst.sh let the program determine specie
# (which will be changed by this incarnation/version of this script)
rMLST_specie = metadataWspecieTsv %>% 
  rename( rMLST_specie = rMLST_result_full_run ) %>%
  select( - `rMLST results` )  # this col was manually created G and has some inconsistency, not using


#rMLST_specie |> View()
```



```{r rMLST_filter_nonEC}
# so still need the rMLST result file to determine what row are E.coli and only keep those. 
# actually, rMLST info in metadataWspecieTsv

# rMLST_specie |> str()
# rMLST_specie %>% View()
rMLST_specie %>% dim()   # 104 rows
rMLST_EC    = rMLST_specie %>% filter( rMLST_specie == "Escherichia coli" )

rMLST_EC %>% dim() #  93 rows ie filtered 11 rows (incl 1 NA)
rMLST_EC %>% colnames()
# "Name" col has the isolate ID that match Gabrielle's "user_sample_name" (these 2 cols seems to match, except 126 missing in user_sample_name)

rMLST_nonEC <- rMLST_specie %>% filter( ! grepl( "Escherichia coli", rMLST_specie   ) )  
rMLST_nonEC %>% dim() #  expect 11 !!   (1 is NA)
rMLST_nonEC %>% print( n=12)

# 11x8
# sample #24 (name=126) has NA
# 10 other are not E.coli

```



filenames eg: AI-1608_103_103_assembly.fasta
need to extract       ^^^
as join key to be able to merge metadata info.
use Regular Expression in str_match() to extract file id number


mlst output examples

mlst   AI-1608_206_206_assembly.fasta
AI-1608_206_206_assembly.fasta  ecoli   10      adk(10) fumC(11)        gyrB(4) icd(8)  mdh(8)  purA(8) recA(2)

mlst  --scheme ecoli AI-1608_206_206_assembly.fasta
AI-1608_206_206_assembly.fasta  ecoli   10      adk(10) fumC(11)        gyrB(4) icd(8)  mdh(8)  purA(8) recA(2)


mlst --legacy --scheme ecoli AI-1608_206_206_assembly.fasta
FILE    SCHEME  ST      adk     fumC    gyrB    icd     mdh     purA    recA
AI-1608_206_206_assembly.fasta  ecoli   10      10      11      4       8       8       8       2



--legacy is easiest to parse
grapetree and phyloviz likely just want the number (for matrix)

/mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv


```{r prep_joinKeys}

mlstResultDF_unfiltered = mlstResultTsv %>% mutate(
  isolateKey = str_match( FILE, "AI[\\-]1608[\\_]([:digit:]+)[\\_]([:digit:]+)_assembly\\.fasta"  )[,2] 
) %>%  rename( forced_EC_scheme = SCHEME )   # to be very clear MLST output were all forced to E.coli

mlstResultDF_unfiltered |> str()


```


```{r join_metadata_and_mlst}

#metadataDF |> str()
metadataDF |> dim()  # 104
metadataDF |> head()  # 104
metadataDF |> select( House_ID, Sample_Type_ID, WGS_ID, WGS_ID_as_string ) |> head()

mlstResultDF_unfiltered |> dim() # 103
mlstResultDF_unfiltered |> head()
mlstResultDF_unfiltered |> select( FILE, forced_EC_scheme, isolateKey ) |>  tail()  # id range: 103..206

meta_and_mlst_DF = full_join(
  x  = metadataDF,
  y  = mlstResultDF_unfiltered, #mlstResultDF,
  by = c( `WGS_ID_as_string` = "isolateKey" )
)


meta_and_mlst_DF %>% head()
meta_and_mlst_DF %>% tail()
meta_and_mlst_DF %>% dim()   # 104... some might not be EC
#meta_and_mlst_DF |> View()   # still have House_ID info here   but Sample_Type_ID and WGS_ID , which as character, has .0 added.  weird!!

meta_and_mlst_DF |> filter( forced_EC_scheme != "ecoli" )  # 0 row, good.

```

```{r join_rMlst}

# filtered rMLST, right join, so non EC would be removed
# expect 93 rows

# meta_and_mlst_DF |> head()
# meta_and_EC_ST |> View()
meta_and_mlst_DF |> dim() # 104x25   ## for a bit was 207 x 25 # not sure why, debugging shouldn't fix it.  stale cache? restarted R then it is ok.
rMLST_EC |> head()
rMLST_EC |> tail()
rMLST_EC |> filter( `MLST results` != "ecoli" )  # Sample 34 Name 136 is not e.coli   
# the `MLST results` col was from metadata gsheet file, that had result from MLST and rMLST pasted side by side to check for non-agreement
# no longer needed by here, will be dropping that col.  (and it is confusing, going by rMLST to specify specie)  #1003  

rMLST_EC |> dim()  # 93



meta_and_EC_ST = right_join(
  x  = meta_and_mlst_DF,
  y  = rMLST_EC,
  #by = c( "WGS_ID" = "Name")
  by = c( "WGS_ID_as_string" = "Name" )
) |> select( - `MLST results` )


meta_and_EC_ST %>% dim()  # 93 rows, good
meta_and_EC_ST %>% colnames() 
#meta_and_EC_ST |>  str()
meta_and_EC_ST |>  head()          # lost House_ID  info.  reset of R fixed it.
meta_and_EC_ST |>  tail()
# meta_and_EC_ST |>  View()       # __key_DF__ __DW_DF__ 

# 

```

# write the combined mlst + metadata file into a new result file
# this was in the old  meta+mlst+grapetree.Rmd, not needed here
```{r save_result, eval=F}


##old previos version## write_tsv( meta_and_mlst_DF, "../Result/meta_and_mlst.tsv" )


## saving new "v2" result // this is optional, in case she wants to look at it.  __key_DF__ __DW_DF__

# write_tsv( meta_and_EC_ST, "../Result/meta_and_EC_ST.tsv" )  
# avoid saving result to public git repo till paper is published
# so save to temporary location and upload to gDrv: https://drive.google.com/drive/u/0/folders/1v-GvAj3N6EZ4LbPpRg1X9DUBhLPn-wFw
write_tsv( meta_and_EC_ST,  r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\Result\joinedRmdDF.tsv)" ) 

```



# massage VFDB table 
simplify, convert to gene present/absent
based on > 80% score as Match.

```{r getFirstNum, eval=T}

## abricate sometime have match value containing semicolon, eg
##  98.75;99.00
##  87.91;100.00;100.00;100.00;100.00
## 100.00;100.00;100.00;100.00             # usda_summary.csv:GCA_052631835.1_PDT002905087.1_genomic_usda.txt
## need to parse them out, maybe keep the first number, up to the first ;

## ref: Env_EcO157/code/stInfo_Snp1.Rmd
## ref: L348, L354.
## ./guatemala_amr/guat_amr_analy.Rmd:## semicolon shows up when multiple match in a given file.
## ./guatemala_amr/guat_amr_analy.Rmd:# many of the genes col could have semicolon separate numbers, eg


# abricate --sumary produce output where
# many of the genes col could have semicolon separate numbers, eg 98.75;99.00
# taking first ; ignore rest, convert to double.
# This kind of output happens when the file has multiple section where it matched.
# running the abricate --db jgrg db/amnt/sequence  will demo this output really well.
# define helper fn that will be used by across() inside a mutate().
getFirstNum = function( charStr ) {
  firstStr = str_match(  charStr , "([:digit:]+\\.*[:digit:]*)")[,1]
  if_else(
    as.numeric(  firstStr ) >= 80 , 1, 0, missing=0     # 80% as cutoff to Positive(1), else Negative(0)
  )
}
```


# join with abricate VFDB result.
# should do this earlier.

```{r massage_vfdb_result, eval=T}
#// vfdb_sum_df_BAK = vfdb_sum_df
#//usdaDB_sum_df = usdaDB_sum_df_text 



VfdbResultDf = VfdbResultTsv     |> 
  rename( file_name = `#FILE` )  |> mutate(
  filename_sid   = str_match( file_name, "AI-1608_([0-9]+)_([0-9]+)_assembly_Abricate_vfdb.tsv")[,2]   ) |>
  mutate( across( 3:length( VfdbResultTsv), getFirstNum ))  

#  mutate( across( 4:length( VfdbResultTsv), getFirstNum ))  
#                 ^-- col idx and length()-1 are very fiddly.  depends on how read* fn worked?.  


VfdbResultDfLng = resfResultTsvLng     |> 
  rename( file_name = `#FILE` )  |> mutate(
  filename_sid   = str_match( file_name, "AI-1608_([0-9]+)_([0-9]+)_assembly.fasta")[,2]   )  # new filename patter (vs wide/summary)
  #no need in Lng ver# mutate( across( 3:length( resfResultTsv), getFirstNum ))  
#VfdbResultDfLng  |> str()  


#- VfdbResultDf_test = VfdbResultDf |> mutate( filename_sid   = str_match( file_name, "AI-1608_([0-9]+)_([0-9]+)_assembly_Abricate_vfdb.tsv")[,2]   )
#- VfdbResultDf_test |> View()
# extract sample ID from file_name, so that it can later be joined to meta data WGS_ID   eg 155
# file_name from vfdb result eg: 
#  AI-1608_155_155_assembly_Abricate_vfdb.tsv
  
VfdbResultTsv |> head(5)
VfdbResultDf  |> head(5)
#VfdbResultDf |> View()
VfdbResultDf  |> str()  
VfdbResultDf  |> select( length(VfdbResultDf)-180 : length(VfdbResultDf) ) |> str()
VfdbResultDf  |> select( 190 : length(VfdbResultDf) ) |> str()
#VfdbResultDf  |> colnames()  
VfdbResultTsv |> dim()    # 103x201
VfdbResultDf  |> dim()    # 103x201
```

# massage Resfinder table 
simplify, convert to gene present/absent
based on > 80% score as Match.
(same code as done for VfdbResultTsv)

```{r massage_resfinder_result, eval=T}
#// resf_sum_df_BAK = resf_sum_df


resfResultDf = resfResultTsv     |> 
  rename( resf_file_name = `#FILE` )  |> mutate(
  filename_sid   = str_match( resf_file_name, "AI-1608_([0-9]+)_([0-9]+)_assembly_Abricate_resfinder.tsv")[,2]   ) |>
  mutate( across( 3:length( resfResultTsv), getFirstNum ))  

resfResultDfLng = resfResultTsvLng     |> 
  rename( resf_file_name = `#FILE` )  |> mutate(
  filename_sid   = str_match( resf_file_name, "AI-1608_([0-9]+)_([0-9]+)_assembly.fasta")[,2]   )   ## diff filename pattern!
  #no need in Lng ver# mutate( across( 3:length( resfResultTsv), getFirstNum ))  



#resfResultDf |> str()
cols2show_resfJoinQC = c( "resf_file_name", "blaCTX", "file", "NUM_FOUND" )
#, "House_ID", "Sample_Type_ID", "WGS_ID", "Isolate_Type", "Source", "ST", "CRO",  "CRO_class_binary", "CTX_M_15", "ST131" )
resfResultDf |> select( matches( cols2show_resfJoinQC ) ) |> str() 

resfResultDfLng |> str() 

```


# looking for source table where CRO/CTX_class_binary came from
(cuz 1 entry had discrepancy)
```{r DBG_looking_for_cro_col, eval=F}

# either of these methods will match colnames with certain regex string
VfdbResultDf |> select( matches("found", ignore.case=T) ) |> colnames()
grep( "found", colnames(VfdbResultDf), value=T, ignore.case=T)

#eg# VfdbResultDf |> select( matches("dra", ignore.case=T) ) |> colnames()


VfdbResultDf |> select( matches("cro|ctx", ignore.case=T) ) |> colnames()
metadataWspecieTsv |> select( matches("cro|ctx", ignore.case=T) ) |> colnames()
# ^^ no cro col from there


meta_and_EC_ST |> select( matches("cro|ctx", ignore.case=T) ) |> colnames()
# CTX_class_binary was found here, later renamed to CRO
metadataDF   |> select( matches("cro|ctx", ignore.case=T) ) |> colnames()
metadataXlsx |> select( matches("cro|ctx", ignore.case=T) ) |> colnames()
# ie from 
#GM_metadata_path =  r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\metadata\GM_metadata.xlsx)"  # weasel
#GM_metadata_path =  "~/dataCache/Borbon_summer_25_Gabrielle/metadata/GM_metadata.xlsx"          # m02 bigmac02
# a series of columns that has Drug_class_binary, so  likely a mistake there, cuz she might have created these col manually.
# stick to the Sample Type info 

```

df/heatmap  plan for GM short paper, CRO, HH, ST, bla.  L404 m02

```{r join_meta_and_EC_ST_andVfdb}
# add VFDB, think of it as abricate result + metadata
# this was done after pathotype skeleton code was added.  the DF used by the pathotype def below, if using it after all, should be updated to use this rather than pathotype_meta ++


meta_and_EC_ST_andVfdb = left_join(   # full_join
  x  = meta_and_EC_ST,
  y  = VfdbResultDf,  
  by = c( `WGS_ID_as_string`  = "filename_sid" )  
  #by = c( "WGS_ID" = "filename_sid" )  
  #by = c( "FILE" = "file_name" )
)    |> rename( CRO_class_binary = CTX_class_binary )   # avoid using this col, likely manually created in metadata xlsx, 1 discrepancy.
#--) |> rename( CRO = CTX_class_binary )

meta_and_EC_ST_andVfdbResf = left_join(     # ++ problem with this join... 
  x  = meta_and_EC_ST_andVfdb,
  y  = resfResultDf,
  by = c( `WGS_ID_as_string`  = "filename_sid" )  
)
```

```{r QC__join_meta_and_EC_ST_andVfdb, eval=F}
meta_and_EC_ST_andVfdb       |> dim()  # 93*236
meta_and_EC_ST_andVfdbResf   |> dim()  # 93*302


cols2show_resfQC = c( "resf_file_name", "file", "WGS_ID", "blaCTX", "House_ID", "Sample_Type_ID", "WGS_ID", "Isolate_Type", "Source", "ST", "CRO",  "CRO_class_binary", "CTX_M_15", "ST131" )

meta_and_EC_ST_andVfdbResf |> select( matches( cols2show_resfQC ))  |> str() # |> head(10)
meta_and_EC_ST_andVfdb     |> select( matches( cols2show_resfQC ))  |> head(10)

resfResultDf |> filter( `blaCTX-M-15_1` != 0 ) |> select( matches( cols2show_resfQC ))  # 11 rows   # filename_sid is NA
#-resfResultDf |> filter( `blaCTX-M-15_1` != 0 ) |> select( matches( cols2show_s2 ))  # 11 rows   # resf_file_name, from renamed #FILE


```

```{r join_meta_and_EC_ST_andVfdb_LONG}

# join table using the Long version of abricate result (instead of wide/summary version)

meta_and_EC_ST_andVfdb_lng = left_join(   # full_join
  x  = meta_and_EC_ST,
  y  = VfdbResultDfLng,  
  by = c( `WGS_ID_as_string`  = "filename_sid" )  
  #by = c( "WGS_ID" = "filename_sid" )  
  #by = c( "FILE" = "file_name" )
)    |> rename( CRO_class_binary = CTX_class_binary )   # avoid using this col, likely manually created in metadata xlsx, 1 discrepancy.
#--) |> rename( CRO = CTX_class_binary )

meta_and_EC_ST_andVfdbResf_lng = left_join(     # ++ problem with this join... 
  x  = meta_and_EC_ST_andVfdb_lng,
  y  = resfResultDfLng,
  by = c( `WGS_ID_as_string`  = "filename_sid" )  
)

meta_and_EC_ST_andVfdbResf_lng |> str()  # 3357*65
#meta_and_EC_ST_andVfdbResf_lng |> View()
```





```{r meta_and_EC_ST_massage}

# massage meta_and_EC_ST so that it is in a more parsable structure for use with heatmap.

#1003 ++   new __DW_DF__ DF for metadata+result
#meta_and_EC_ST_parsable =  meta_and_EC_ST |>
#meta_and_EC_ST_andVfdb_parsable_step1 =  meta_and_EC_ST_andVfdbResf |>
meta_ST_Db_step1 =  meta_and_EC_ST_andVfdbResf |>
    mutate(
    Source = case_when( 
      str_detect( Isolate_Type  , regex("water", ignore_case=T)  )  ~ "Water", 
      str_detect( Isolate_Type  , regex("mother", ignore_case=T)  )  ~ "Mother", 
      str_detect( Isolate_Type  , regex("Child", ignore_case=T)  )  ~ "Child",
      TRUE ~ "FIXME"
      ),
    CRO_meta = case_when(   # ie CRO info from the metadata sheet, pulled from Isolate_Type info.  
      str_detect( Isolate_Type  , regex("Ceftriaxone", ignore_case=T)  )  ~ "CRO+",   # changed to binary/double for now, to compare with CRO col
      TRUE ~ "CRO-"
    ))  |> rename( CRO = CRO_meta )

# add_cols_heatmap_cols_from_ST_info
#++ placed ST in the x-axis of the heatmap, so the ST cols added by this Step2 currently not utilized #1003.
meta_ST_Db_step2 = meta_ST_Db_step1 |>  mutate (
  ST131 = case_when( ST == "131" ~ "Yes" , TRUE ~ "No" ),
  ST10  = case_when( ST ==  "10" ~ "Yes" , TRUE ~ "No" ),
  ST69  = case_when( ST ==  "69" ~ "Yes" , TRUE ~ "No" ),  
)
## there are like 44 ST, just pick the important ones? ++ 


```

```{r QC_sanity_check_data, eval=F}

meta_ST_Db_step2 |> group_by( CRO ) |> summarize( count = n() )

# check data records count make sense.
# eg 37 CRO+ and 56 CRO-
# okay cuz not every sample was has CRO resistance.  (but would be strange if mom vs child differ...)

# check every household has mom/child/water

meta_ST_Db_step2 |> group_by( Source ) |> summarize( count = n() )

meta_ST_Db_step2 |> group_by( House_ID ) |> summarize( count = n() )  # some house has multiple child? eg SD0008 has n=6  ++

```

```{r meta_and_EC_ST_massage_LONG}

#1004  here version is using LONG version of Vfdb and Restfinder results
# should not have the various ST and CTX (wide) cols
# though the metadata from GM still have things like CZ_class_binary,
# so not quite true "tidy"

#-- lng_meta_ST_Db_step1 =  meta_and_EC_ST_andVfdbResf_lng |>
meta_ST_Db_lng =  meta_and_EC_ST_andVfdbResf_lng |>
    mutate(
    Source = case_when( 
      str_detect( Isolate_Type  , regex("water", ignore_case=T)  )  ~ "Water", 
      str_detect( Isolate_Type  , regex("mother", ignore_case=T)  )  ~ "Mother", 
      str_detect( Isolate_Type  , regex("Child", ignore_case=T)  )  ~ "Child",
      TRUE ~ "FIXME"
      ),
    CRO_meta = case_when(   # ie CRO info from the metadata sheet, pulled from Isolate_Type info.  
      str_detect( Isolate_Type  , regex("Ceftriaxone", ignore_case=T)  )  ~ "CRO+",   # changed to binary/double for now, to compare with CRO col
      TRUE ~ "CRO-"
    ))  |> rename( CRO = CRO_meta )

#^^ contrast to  meta_ST_Db (below) #1004  new __DW_DF__ DF for metadata+result  LONG format (instead of pivot_longer)
```




```{r coding_aid_resfinder_result, eval=F}
cols2show_s2 = c( "ID", "resf_file_name", "WGS", "blaCTX", "House_ID", "Sample_Type_ID", "WGS_ID", "Isolate_Type", "Source", "ST", "CRO",  "CRO_class_binary", "CTX_M_15", "ST131" )


meta_ST_Db_step2 |> filter( ST131 == "Yes" )    # only 2
mlstResultDF_unfiltered  |> filter( ST=="131")  # only 2 rows indeed  ++ ponder on resulting figure ++

resfResultDf |> filter( `blaCTX-M-15_1` != 0 )  # 11 rows
resfResultDf |> filter( `blaCTX-M-15_1` != 0 ) |> select( matches( cols2show_s2 ))  # 11 rows
#resfResultDf |> filter( `blaCTX-M-15_1` != 0 ) |> select( any_of( cols2show_s2 ))  # 11 rows



#meta_and_EC_ST_andVfdb_parsable_step1 |> str()
#meta_and_EC_ST_andVfdb_parsable_step1 |> select( matches("bla", ignore.case=T) ) |> colnames()
#meta_ST_Db_step1 |> select( matches("bla", ignore.case=T) ) |> colnames()
# VfdbResultDf |> select( matches("c", ignore.case=T) ) |> colnames()
meta_ST_Db_step1 |> select( matches("st", ignore.case=T) ) |> str() #colnames()

meta_ST_Db_step2 |> select( matches("ctx", ignore.case=T) ) |> str() #colnames()
# blaCTX-M-15_1  is the only 15*   ## fix NA++ 

```

```{r add_cols_heatmap_cols_from_vfg_anal }

#1003 new __DW_DF__ DF for metadata+result

# add_cols_heatmap_cols_from_resfiner_anal
meta_ST_Db_step3 = meta_ST_Db_step2 |>  mutate (
  CTX_M_15 = if_else(  rowSums( pick( matches( "blaCTX")), na.rm=T) >= 1 , 1, 0, missing=-1)
)
#++ ^^ actually, above likely isn't good approach.  
#better: 
# imp_resist_gene = "blaCTX-M-15", "blaTEM-1", "blaTHV-13"  ## ie actual text in cell, just like ST col has various numbers in it. ++ 
# actually instead of pivot_longer, can use the long version of abricate-resfind result (instead of summary)

cols2show_s3 = c( "House_ID", "Sample_Type_ID", "WGS_ID", "Isolate_Type", "Source", "ST", "CRO",  "CRO_class_binary", "CTX_M_15", "ST131" )
#meta_ST_Db_step3 |> str()
meta_ST_Db_step3 |> filter( CTX_M_15 >= 1 )  # 27 now
meta_ST_Db_step3 |> filter( CTX_M_15 >= 1 ) |> select( matches( cols2show_s3 )) 
# meta_ST_Db_step3 |> select( matches( cols2show_s3 )) |> head(10) 
# ++ ^^ might be okay, better at check#2 after fixing NA in SID matching problem;  more check as pass 3 TBD ++


meta_ST_Db_step4 = meta_ST_Db_step3 |>  mutate (
  Block = case_when(
    (ST131 == "Yes" & CTX_M_15 >= 1)  ~ "Block 1",
    TRUE ~ "Block 0"  
  )) 


meta_ST_Db_step4 |> group_by( Block ) |> summarize( freq = n() )
meta_ST_Db_step4 |> filter( Block == "Block 1" ) |> select( matches( cols2show_s3 ))  # WGS_ID 182, SID 4
# Block 1 only has 1 entry!  ++ ponder resulting fig ++

#--meta_and_EC_ST_andVfdb_parsable = meta_and_EC_ST_andVfdb_parsable_step2  
meta_ST_Db = meta_ST_Db_step4   #1003 new __DW_DF__ DF for metadata+result

#^^ contrast to  lng_meta_ST_Db  #1004
```

(bmail 2026.1004 ...  TBD)  ++
Blocks... You wanted to define Block 1 as "ST131 and CTX-M-15"? 
if so, there is only 1 row matching this, for House_ID 38
​The code for picking this out is tricky thus not done in this draft yet, 
but also wanted to discuss this Block idea further first.


```{r df4heatmap_ST}

#++ maybe almost ready for df4heatmap_ST and another df4heatmap_bla ++
#1003 ++ Continue here ^^    

meta_ST_Db |> str()  #1003 new __DW_DF__ DF for metadata+result

meta_ST_Db |> select( House_ID, Sample, WGS_ID_as_string, Source, ST131, CTX_M_15, Block ) |> str()

#cols4heatmap_ST = c( "House_ID", WGS_ID_ )
#  select( any_of(  cols4heatmap_ST ))

cols_ST = c( "ST131", "ST10", "ST" ) 

df4heatmap_ST = meta_ST_Db |>
  select( House_ID, Sample, WGS_ID_as_string, Source, ST131, CTX_M_15, Block, cols_ST )

fig_heatmap_ST = 
ggplot( df4heatmap_ST,
        aes( y = House_ID, x = ST  )) +
  geom_tile( fill="royalblue1"  ) +
  facet_wrap( ~ factor(Source, levels=c("Mother", "Child", "Water") )  )+ 
  #facet_wrap( ~ Block ) + 
  #facet_wrap( ~ Source + Block  ) +   ## kinda works, but still messy, only 1 row, don't need this.
  labs( 
    title = "ST comparison by Source and Household",
  ) + 
  theme_minimal() +
  theme(
    axis.text.x = element_text( angle=90 )
  )

fig_heatmap_ST 

## only 1 House_ID was CTX_M_15, just highlight that row?
## else, might have to create a new col {Block}+{House_ID}

```
```{r fig_long_df4heatmap_Bla}

# using the resfinder data for high resistance genes

# use meta_ST_Db_lng

#meta_ST_Db_lng |> View() 
meta_ST_Db_lng |> str() 

df4heatmap_Resf = meta_ST_Db_lng |>
  select( House_ID, Sample, WGS_ID_as_string, GENE.y, Source, CRO )  # Block? 

df4heatmap_ResfBla = meta_ST_Db_lng |>
  select( House_ID, Sample, WGS_ID_as_string, GENE.y, Source )  |>
  filter( str_detect( GENE.y, regex("bla") ))


fig_heatmap_Bla = 
ggplot( df4heatmap_ResfBla,
        aes( y = House_ID, x = GENE.y, 
             #fill=CRO  
             )) +
  #geom_tile(  ) +
  geom_tile( fill="gold3"  ) +
  facet_wrap( ~ factor(Source, levels=c("Mother", "Child", "Water") )  )+ 
  #facet_wrap( ~ Source ) + 
  #facet_wrap( ~ Source + Block  ) + 
  labs( 
    title = "BLA Resistance Gene comparison by Source and Household // fill=NADA ",
  ) + 
  theme_minimal() +
  theme(
    axis.text.x = element_text( angle=90 )
  )

fig_heatmap_Bla 
# color by 3GC plate ?  CRO+ vs CRO- ?   Fill=CRO didn't change anything.  ++  



```

```{r QC_check_CRO_col, eval=F}   

# see DBG_looking_for_cro_col
# CRO came from col GM_metadata.xlsx col CTX_class_binary
# likely manually created by her, so sticking to Sample_Type col info parsed programatically 

#meta_and_EC_ST_andVfdb_parsable |> select( CRO_meta, CRO_class_binary ) |> head() #str()
#meta_and_EC_ST_andVfdb_parsable |> select( CRO, CRO_class_binary ) |> head() #str()
meta_ST_Db |> select( CRO, CRO_class_binary ) |> head() #str()

#--QC_meta_and_EC_ST_andVfdb_parsable = meta_and_EC_ST_andVfdb_parsable |> mutate(
QC_meta_and_EC_ST_andVfdb_parsable = meta_ST_Db |> mutate(
    CRO_meta_asNum = case_when(   # ie CRO info from the metadata sheet, pulled from Isolate_Type info.  
      str_detect( Isolate_Type  , regex("Ceftriaxone", ignore_case=T)  )  ~ 1, # "CRO+",   # use binary/double, to compare with CRO col
      TRUE ~ 0 # "CRO-"
  )) |> mutate( 
    CRO_agreement = case_when(
      CRO_meta_asNum == CRO_meta_asNum ~ "concordance",
      TRUE ~ "CHECK_RESULT"
    ))
QC_meta_and_EC_ST_andVfdb_parsable |> group_by( CRO_agreement ) |> summarize( freq=n() )

QC_meta_and_EC_ST_andVfdb_parsable |> filter( CRO_agreement == "CHECK_RESULT" )

QCcols2show = c( "House_ID", "Sample_Type_ID", "WGS_ID", "Isolate_Type", "Source", "ST", "CRO_meta", "CRO_meta_asNum",  "CRO" )

QC_meta_and_EC_ST_andVfdb_parsable |> filter( CRO_agreement == "CHECK_RESULT" ) |> select( any_of( QCcols2show ))

# earlier had 2 row NOT in agreement... cuz CRO col from metadata likely had error.  sticking with info parsed out of Sample_Type_ID. ## resolved.

```

```{r background_info_for_heatmap_code_creation, eval=F}

# QA/debug use, thus eval=F   won't have real need after code is written

cols2show = c( "House_ID", "Sample_Type_ID", "WGS_ID", "Isolate_Type", "Source", "ST", "CRO",  "CRO_class_binary" )  # CRO_meta 


#--meta_and_EC_ST_andVfdb_parsable |> filter( Source == "FIXME" )  # when did full_join with VFDB, 10 rows has no matching metadata ## 0 rows when left_join used and all VFDB result has metadata info
meta_ST_Db |> filter( Source == "FIXME" )  # when did full_join with VFDB, 10 rows has no matching metadata ## 0 rows when left_join used and all VFDB result has metadata info

meta_ST_Db |> select( any_of( cols2show) ) |> head()
meta_ST_Db |> group_by( Source ) |> summarize( freq = n() ) |> janitor::adorn_totals("row")

#old,before shuffle# meta_and_EC_ST_parsable |> group_by( CRO ) |> summarize( freq = n() ) |> janitor::adorn_totals("row")
meta_ST_Db |> group_by( CRO_meta ) |> summarize( freq = n() ) |> janitor::adorn_totals("row")

# meta_and_EC_ST_parsable |> group_by( ST ) |> summarize( freq = n() ) |> janitor::adorn_totals("row")
meta_ST_Db |> select( ST ) |> unique()  # 43 ST and "-"   

# heatmap need (maybe in tidy format)
# Source: Mom, Child, Water
# CRO: 1 or 0 : 3GC plate vs non-selective media.  // CRO is acronym for ceftriaxone
# ST
# Household
# Block:   1, or 2  where 1=High Sharing, 2=Only Human ST (compute this based on ST131, CTX-M-15... )

# Num of Genes ?


# another heatmap
# 1/0 for presence of blaCTX-M-15, -14, blaTEM-1, blaTHV-... blaSHV... maybe list all of them? 

## determine number of ST in dataset 43 + "-"
## determine number of bla* genes


```


########################################## =====================================
########################################## =====================================

// rest not useful now for 10.03 new heatmap generation for short paper (PLOS)

########################################## =====================================
########################################## =====================================

# See irrExp26 for updated definition and code for pathotype , code below for pathotype table need update if decides to use it ++


```{r determinePathotype}

pathotype_names = c( "ExPEC", "ETEC", "EPEC", "STEC", "EAEC", "DAEC", "EIEC" )  # "EIEC_Shigella",  "STEC_EHEC"

ETEC_genes = c( "lt", "sta" )
EPEC_genes = c( "eae" )         # this dataset has only eae
STEC_genes = c( "stx" )         # no stx* in vfdb result
EIEC_genes = c( "ipaH" )        # no ipa* in vfdb result
EAEC_genes = c( "aatA", "aggR", "aaiC" )        # OR condition, none exist in vfdb result
#DAEC_genes = c( "afa", "dra", "daa" )           # OR condition, names in VFDB result are more specific
DAEC_genes = c( "afaA", "afaB-I", "afaC-I", "afaD", "afaE-II", "afaF-III", "draA", "draB", "draC", "draD", "draE2", "draP", "daaD", "daaE", "daaF" )

# most gene definition from 
# doi:10.1097/QCO.0000000000000665   // my annotated file: file:///J:/Shared%20drives/Tin_Ho_gDrv_BL_shareDrv/mph_hw/mph_prj_bioinfo/ExPEC_Ala/Defining%20diarrheagenic%20E.%20coli.pdf 
# Table 1 p 16 
# DAEC said: Genes in the afa, dra, or daa adhesin operons; cross-reactivity with EAEC genes  (and that more work needed for better definition).  so just say predicted possible pathotype.
# Table 2 p 17, paper reported afaE-I or afaE-III


ExPEC_genes = c( "papA","papA","papB", "papC","papD","papE","papF","papG","papH" , "sfaC","sfaY", "kpsMII", "iutA" )   # only kpsM, no kpsMII

#++# consider meta_and_EC_ST_andVfdb_parsable instead of VfdbResultDf  if ever get pathotype code completed (ref irrExp26)
pathotypeDf = VfdbResultDf |>
  mutate(
    #ExPEC = if_else( papA + papC + sfaC + kpsMII + iutA >= 2, "MoreWorkNeeded", 0, missing=0 ),   # ExPEC definition: at least 2 of the 5 genes
    #ETEC  = if_else( lt + sta >= 1, "lt/ltA check", 0, missing=0 ),   # ETEC definition: at least 1 of the 2 genes   # no lt in this dataset
    #ETEC  = if_else(  if_any(any_of("lt"))  + if_any(any_of("sta")) >= 1, "lt/ltA check", 0, missing=0 ),   # ETEC definition: at least 1 of the 2 genes   # no lt in this dataset
    ExPEC = if_else(  rowSums( pick( any_of( ExPEC_genes)), na.rm=T  ) >= 2, 0.9, 0, missing=-1 ),     # ExPEC definition: at least 2 of 5 genes, and more nuanced... more work needed ++
    ETEC  = if_else(  rowSums( pick( any_of( ETEC_genes )), na.rm=T  ) >= 2, 1, 0, missing=-1 ),   # ETEC definition: at least 1 of the 2 genes   # this seems to work.  no lt or sta in this dataset.
    EPEC  = if_else(  rowSums( pick( any_of( EPEC_genes )), na.rm=T  ) >= 1, 1, 0, missing=-1 ),
    STEC  = if_else(  rowSums( pick( any_of( STEC_genes )), na.rm=T  ) >= 1, 1, 0, missing=-1 ),
    EIEC  = if_else(  rowSums( pick( any_of( EIEC_genes )), na.rm=T  ) >= 1, 1, 0, missing=-1 ),
    EAEC  = if_else(  rowSums( pick( any_of( EAEC_genes )), na.rm=T  ) >= 1, 1, 0, missing=-1 ),   # or condition of 3 genes, thus still just checking for >=1 
    DAEC  = if_else(  rowSums( pick( any_of( DAEC_genes )), na.rm=T  ) >= 1, 1, 0, missing=-1 )    # or condition of 3 genes, thus still just checking for >=1 
    )

#pathotypeDf |> select( file_name, ETEC  )  |> head(10)
#pathotypeDf |> filter( ETEC != "No" ) |> select( file_name, ETEC  )  |> head(10)  # 0 rows
#pathotypeDf |> filter( EPEC != "No" ) |> select( file_name, EPEC  )  |> head(10)  # 1 rows

pathotypeDf |> filter( if_any( any_of( pathotype_names ), ~ .x != 0 )) |> select( file_name, filename_sid, any_of( pathotype_names)  )  |> head(25)  # 21 rows

#pathotypeDf |> View()

# while I was looking at the paper, copilot came up with the above.  i was going to use case when.  the gene names are wrong.  the ExPEC logic... did i use it before?
# search email ...
# https://doi.org/10.1038/s41579-020-0416-x  // dont look like local def
# Saba had a UPEC Local and Global... manuscript (not out yet) , it states: 
# study, a different molecular definition by Johnson et al. for the broader ExPEC group (presence
# of ≥2 of papAH and/or papC, sfa/focDE, afa/dra, iutA, kpsMII) was the strongest predictor of
# experimental virulence in a murine model.

```


# save pathotypeDf   and color cell in spreadsheet to highlight the yes
# include Ceftriaxone status   so can just use spreadsheet to group/filter.  not too many records.



```{r pathotype_meta_join}

# consider meta_and_EC_ST_andVfdb_parsable instead of meta_and_EC_ST  if ever get pathotype code completed (ref irrExp26)
# meta_and_EC_ST_andVfdb_parsable has more parsable metadata, but then VFDB result may get duplicated... 
pathotype_meta = full_join(
  x  = meta_and_EC_ST,
  y  = pathotypeDf,  
  by = c( `WGS_ID_as_string`  = "filename_sid" )  
  #by = c( "WGS_ID" = "filename_sid" )  
  #by = c( "FILE" = "file_name" )
) |> rename( CRO = CTX_class_binary )

  # |> mutate(
  # CRO = case_when( 
  #  grepl( "Ceftriaxone", Isolate_Type, ignore.case=T ) ~ "Positive",
  #  TRUE ~ "Negative" ) )
  # already have CTX_class_binary

# file_name from vfdb result eg: AI-1608_155_155_assembly_Abricate_vfdb.tsv
# FILE                eg AI-1608_103_103_assembly.fasta
# file_name           eg 
# fasta_used_by_rMLST eg AI-1608_103_103_assembly

# pathotype_meta |>  View()  # ~10 rows have NA, check with Gabrielle which NA are expected, which rows to filter, what to keep.



# sanity check

meta_and_EC_ST |> colnames()

#pathotype_meta |> select( file_name, WGS_ID_as_string, any_of( pathotype_names) ) |> head(25)
pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, any_of( pathotype_names), CRO ) |> head(25)
#pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, forced_EC_scheme, `MLST results`, rMLST_specie, forced_EC_scheme, ST, any_of( pathotype_names), CRO ) |> View()
# pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, any_of( pathotype_names), CRO ) |> View()

#1003 `MLST results` col removed, cuz it was creating confusion in the data. 
pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, forced_EC_scheme, rMLST_specie, forced_EC_scheme, ST, any_of( pathotype_names), CRO ) |> filter( is.na( House_ID))
# why these 10 rows has NA ? 
# 93 rows, remaining 10 don't have ST, rMLST_specie info, etc.

```


```{r write_summary_pathotype_result, eval=F}

sum_pathotype_result_brief = pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, any_of( pathotype_names), CRO ) 

sum_pathotype_result = pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, forced_EC_scheme, `MLST results`, rMLST_specie, forced_EC_scheme, ST, any_of( pathotype_names), CRO ) 


write_tsv( sum_pathotype_result, "../Result/pathotype.tsv" )

#0614 draft result.  ExPEC need logic update.  rest of pathotype need review with Jay on logic being acceptable. ++



```


gDrive for Borbon data, pathotype result uploaded there
https://drive.google.com/drive/u/0/folders/1eDXqacdExuWbvYvnBdqHIjb1QMkfTAJZ


have more columns from metadata and VFG that can be included as needed

Methods write up:


In-silico prediction of pathotype.  
Abricate is run on the WGS of the isolates to determine the virulence genes each isolate carries.
Genes that predict Pathotypes as described in Jesser and Levy (2020) [ doi:10.1097/QCO.0000000000000665 ] is screened for.
Briefly, 
presence of both _lt_ and _sta_ genes would be predictive of ETEC pathotype (none in this dataset).  
Presence of _eae_ genes would be predictive of EPEC pathotype (1 isolate in this dataset).
Presence of _stx_ genes would be preditive of STEC pathotype (none in this dataset).
Presence of _ipaH_ genes would be preditive of EIEC pathotype (none in this dataset).
Presence of any of _aatA_, _aggR_ or _aaiC_, genes would be preditive of EAEC pathotype (none in this dataset).
Presence of any of _afa_, _dra_ or _daa_ family of genes would be preditive of DAEC pathotype (~6 this dataset).

For ExPEC, we used a local definition (paper for reference) [More work TBD]

Please note that the pathotype is predicted from computational screening of DNA sequences, and, as stated by Jesser and Levy, some of these molecular based pathotype definition can use of further refinements.  Thus, please use these predicted pathotypes with caution.



(Figure caption, or even table headings, should emphasize it is predicted pathotype.)




```{r pathotype_prevalence, eval=T}
# Need a table as agreed, which have counts (and %)
# 

# also, check that 93 rows with data is the correct count ++


### make a smaller 2D_table first
Patho_Source_DF = pathotype_meta |> select( House_ID, WGS_ID_as_string, source_mother, source_child, source_water, rMLST_specie, any_of( pathotype_names), CRO ) 
#Patho_Source_DF |> View()

Patho_Source_DF_long = Patho_Source_DF |> pivot_longer( pathotype_names, names_to="Pathotype", values_to="Prevalence" )
# Patho_Source_DF_long |> View()

PathoPrevChild  = Patho_Source_DF_long |> filter( source_child  == 1 ) |> group_by( Pathotype ) |> summarize( Child_Prev = sum(Prevalence))
PathoPrevMother = Patho_Source_DF_long |> filter( source_mother == 1 ) |> group_by( Pathotype ) |> summarize( Mother_Prev = sum(Prevalence))
PathoPrevWater  = Patho_Source_DF_long |> filter( source_water  == 1 ) |> group_by( Pathotype ) |> summarize( Water_Prev = sum(Prevalence))
# 16 MaybeExPEC, 2 from water , 7 each mother and child, add up right
# above split in 3, if concat, should be the right result

# PathoPrevChild

PathoPrevTemp = left_join(
  x  = PathoPrevMother, 
  y  = PathoPrevChild
)

PathoPrevAll = left_join(
  x  = PathoPrevTemp, 
  y  = PathoPrevWater
)

PathoPrevAll    #0614   this table decent, pasted to gSheet for discussion
# https://docs.google.com/spreadsheets/d/1sxljIo1sj4x9fDSP2lfo0vXg6DMgh8vkq833sdX3-UE/edit?gid=1178728147#gid=1178728147


# ++ OH... Need to check denominator count.  E.coli?  what rows had been filtered?  ++

# if desire to find prevalence count by CRO +ve vs -ve, 
# do a Patho_Source_DF_long |> filter( CRO ... ) 
# and recreate the table.

```

```{r double_pivot_longer_create_problem__dont_use, eval=F}
Patho_Source_DF_long_src = Patho_Source_DF_long |> pivot_longer( c(`source_mother`, `source_child`, `source_water` ), names_to="Source", values_to="Source_count" ) # dont like name...
#^^ this second pivot may be messing with values.  creating extra counts... thus the problem.
#^^ instead of creating 3 cols for source_mother, source_child, source_water, create a single col with 3 names, then don't have to pivot twice.  #0614



Patho_Source_DF_long_src  |> group_by( Pathotype, Source ) |> summarize( PathoPrev = sum(Prevalence))
#^^ result is wrong this way!!   sum() didn't split by the 2nd group_by... 

# deleting rest of pathotype prevalence count code that didn't produce correct result, 
# cuz the 2nd pivot_long induced extra counts and corrupt the process
# if needed, see git log e7b7ca2  for the old code
```


```{bash}
date
```
```{r date}
#Sys.Date()
Sys.time()
```

```{r litmus_test, eval=F}
#print( THE_END )   # expect error here :D

```


