---
title: "generate Pathotype table (Borbon Summer 2025)"
date: "2026-06-06"
output:
  html_document: default
  pdf_document: default
---


(code based on meta+mlst+grapetree.Rmd)

Generate pathotype table for Gabrielle's Borbon Summer 2025 dataset
(Jay suggestion bmail 2026.06~05)

dataCache: 
/mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT 

use Abricate_vfdb_summary.csv
don't use mlst.all.tsv ?   but the xlsx used below (for grapetree generation also).

eg desired result table:

Pathotype\source  mom  |  child   |  water
---------  ------------|----------|---------
ExPEC           n (%)  |  n (%)   |  n (%)
ETEC            n (%)  |  n (%)   |  n (%)
EPEC            n (%)  |  n (%)   |  n (%)
EAEC            n (%)  |  n (%)   |  n (%)
DAEC            n (%)  |  n (%)   |  n (%)
EIEC/Shigella   n (%)  |  n (%)   |  n (%)  
STEC/EHEC       n (%)  |  n (%)   |  n (%)      # think STEC would be 0



Does it matter if Ceftriaxone +ve or -ve (CRO+ vs CRO-)? 
Separate table for the two media?
A> Combined for manuscript, separated for self evaluation and highlight interesting points in discussion.  maybe as supplements?



Gabrielle, whatever write up that you already have, if you can send me a copy, that would help me get more context and remember where we left off :D 


inputs:
- meta data for sample source, CRO+ vs CRO- (from "Sample Type" col in the xlsx file, but need to extract CRO+ vs CRO- info from the string),
- abricate vfg (not resfinder) result   for the genes used to define Pathotype.
  https://drive.google.com/file/d/1DD_T15Uz0C_zydz8EeNN8fbX_MPqG5__/view?usp=drive_link 
  /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv    104 rows

~~~~

ExPEC definition.  Use local def from Liseth.  citation?

other Pathotype def: 
Paper by Kelsey J. Jesser, Karen Levy: 
Curr Opin Infect Dis. 2020 October ; 33(5): 372–380. doi:10.1097/QCO.0000000000000665

My annotated copy:
J:\Shared drives\Tin_Ho_gDrv_BL_shareDrv\mph_hw\mph_prj_bioinfo\ExPEC_Ala\ Defining diarrheagenic E. coli.pdf 
https://drive.google.com/drive/u/0/folders/1nKbMtrcIxkGjHMCUsdoQvPtEL1MiLbPS


~~~

gDrive for Borbon_summer_25:
https://drive.google.com/drive/u/0/folders/1eDXqacdExuWbvYvnBdqHIjb1QMkfTAJZ


(
previously 
ran rMLST to determine organism, non E.coli will be fitlered out.  result of rMLST placed in new version of metadata file
GM all samples + species 
https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
[saved to metadata/GM_all_samples+species.tsv]
)



~~~



```{r setup}

library(pacman)
p_load(tidyverse)
p_load( dplyr )
p_load( readr )
p_load( stringr )


```


```{bash}

# path for MLST result
pwd
ls -la /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/metadata

ls -la /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv    
# 104 rows

```


# read inputs

```{r read_inputs}


# metadata info
# still using this metadata file, need to extract CFX+ vs CFX- from "Sample Type" col
metadataXlsx = readxl::read_xlsx( r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\metadata\GM_metadata.xlsx)", sheet="Sample IDs + Susceptibility", na="" )  # enclose windows path with r"(...)"  # strange syntax, but whatever.
metadataXlsx %>% dim()  
#metadataXlsx %>% View()
#metadataXlsx %>% head(5)
metadataXlsx %>% colnames()


#VfdbResultFile = "/mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv"
VfdbResultFile = "c:/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv"          # weasel
VfdbResultTsv  = read_tsv( VfdbResultFile, skip=0, show_col_types=F )  



VfdbResultTsv |> head(5) 
VfdbResultTsv |> colnames()
VfdbResultTsv |> dim()      # 103 rows, 201 cols


# specie info from rMLST 
# from MLST, and some are not e.coli.  determination of e.coli will now be determined by rMLST, 
# with info manually merged into the metadata file 
# GM all samples + species 
# https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
# mostly treat this file for rMLST specie determination using column rMLST_result_full_run
# the `MLST results` was determined when run_mlst.sh let the program determine specie
# (which will be changed by this incarnation/version of this script)
metadataWspecieTsv = read_tsv( "../metadata/GM_all_samples+species.tsv" )
metadataWspecieTsv %>% colnames()  
metadataWspecieTsv %>% dim()       # 104 x 9

```



```{note eval=F}

# run rMLST on fasta that MLST said is not ecoli

```

```{r massage_input_file, eval=T}

# metadata, WgS_ID is saved as number, need it as string to join
metadataDF = metadataXlsx %>% 
  mutate( WGS_ID_as_string = as.character(WGS_ID) ) %>% 
  rename( Sample_Type_ID = Sample_Type ) %>% 
  rename( Isolate_Type = 'Sample Type' )


# GM all samples + species 
# https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
# mostly treat this file for rMLST specie determination using column rMLST_result_full_run
# the `MLST results` was determined when run_mlst.sh let the program determine specie
# (which will be changed by this incarnation/version of this script)
rMLST_specie = metadataWspecieTsv %>% 
  rename( rMLST_specie = rMLST_result_full_run ) %>%
  select( - `rMLST results` )  # this col was manually created G and has some inconsistency, not using


#rMLST_specie |> View()
```



```{r rMLST_filter_nonEC}
# so still need the rMLST result file to determine what row are E.coli and only keep those. ++
# actually, rMLST info in metadataWspecieTsv

# rMLST_specie %>% View()
rMLST_specie %>% dim()   # 104 rows
rMLST_EC    = rMLST_specie %>% filter( rMLST_specie == "Escherichia coli" )

rMLST_EC %>% dim() #  93 rows ie filtered 11 rows (incl 1 NA)
rMLST_EC %>% colnames()
# "Name" col has the isolate ID that match Gabrielle's "user_sample_name" (these 2 cols seems to match, except 126 missing in user_sample_name)

rMLST_nonEC <- rMLST_specie %>% filter( ! grepl( "Escherichia coli", rMLST_specie   ) )  
rMLST_nonEC %>% dim() #  expect 11 !!   (1 is NA)
rMLST_nonEC %>% print( n=12)


```



filenames eg: AI-1608_103_103_assembly.fasta
need to extract       ^^^
as join key to be able to merge metadata info.
use Regular Expression in str_match() to extract file id number


mlst output examples

mlst   AI-1608_206_206_assembly.fasta
AI-1608_206_206_assembly.fasta  ecoli   10      adk(10) fumC(11)        gyrB(4) icd(8)  mdh(8)  purA(8) recA(2)

mlst  --scheme ecoli AI-1608_206_206_assembly.fasta
AI-1608_206_206_assembly.fasta  ecoli   10      adk(10) fumC(11)        gyrB(4) icd(8)  mdh(8)  purA(8) recA(2)


mlst --legacy --scheme ecoli AI-1608_206_206_assembly.fasta
FILE    SCHEME  ST      adk     fumC    gyrB    icd     mdh     purA    recA
AI-1608_206_206_assembly.fasta  ecoli   10      10      11      4       8       8       8       2



--legacy is easiest to parse
grapetree and phyloviz likely just want the number (for matrix)

/mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv


```{r prep_joinKeys}

mlstResultDF_unfiltered = mlstResultTsv %>% mutate(
  isolateKey = str_match( FILE, "AI[\\-]1608[\\_]([:digit:]+)[\\_]([:digit:]+)_assembly\\.fasta"  )[,2] 
) %>%  rename( forced_EC_scheme = SCHEME )   # to be very clear MLST output were all forced to E.coli

```


```{r join_metadata_and_mlst}

meta_and_mlst_DF = full_join(
  x  = metadataDF,
  y  = mlstResultDF_unfiltered, #mlstResultDF,
  by = c( `WGS_ID_as_string` = "isolateKey" )
)


meta_and_mlst_DF %>% head()
meta_and_mlst_DF %>% dim()   # 104... some might not be EC
meta_and_mlst_DF |> View()

#0613 zzz
```

```{r join_rMlst}

# filtered rMLST, right join, so non EC would be removed
# expect 93 rows

meta_and_EC_ST = right_join(
  x  = meta_and_mlst_DF,
  y  = rMLST_EC,
  by = c( "WGS_ID" = "Name")
)

meta_and_EC_ST %>% dim()  # 93 rows, good
meta_and_EC_ST %>% colnames() 
```

# write the combined mlst + metadata file into a new result file
```{r save_result}

##old previos version## write_tsv( meta_and_mlst_DF, "../Result/meta_and_mlst.tsv" )


## saving new "v2" result // this is optional, in case she wants to look at it.  __key_DF__ __DW_DF__

# write_tsv( meta_and_EC_ST, "../Result/meta_and_EC_ST.tsv" )  
# avoid saving result to public git repo till paper is published
# so save to temporary location and upload to gDrv: https://drive.google.com/drive/u/0/folders/1v-GvAj3N6EZ4LbPpRg1X9DUBhLPn-wFw
write_tsv( meta_and_EC_ST,  r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\Result\joinedRmdDF.tsv)" ) 

```

## split CFX+ vs CFX-  


```{r split_cfx_pos_vs_neg}


meta_and_EC_ST__CfxPos = meta_and_EC_ST %>%
  filter(   grepl( "Ceftriaxone", Isolate_Type, ignore.case=T ) ) 

meta_and_EC_ST__CfxNeg = meta_and_EC_ST %>%
  filter( ! grepl( "Ceftriaxone", Isolate_Type, ignore.case=T ) ) 


meta_and_EC_ST__CfxPos %>% dim()  # 37
meta_and_EC_ST__CfxNeg %>% dim()  # 56

# meta_and_EC_ST__CfxNeg %>% View()
```

```{r prepGrapeTreeDf}

# Ceftriaxone Positive

grapeTree_1prof_Borbon_CfxPos = meta_and_EC_ST__CfxPos %>% select( ST, adk,	fumC,	gyrB,	icd,	mdh,	purA,	rec ) %>%
  rename(`#Name` = "ST" ) 
  
# metadata
grapeTree_2aux_Borbon_CfxPos = meta_and_EC_ST__CfxPos %>% select( ST, Isolate_Type, rMLST_specie ) %>%
  rename( `id` = "ST")

# Ceftriaxone Negative

grapeTree_1prof_Borbon_CfxNeg = meta_and_EC_ST__CfxNeg %>% select( ST, adk,	fumC,	gyrB,	icd,	mdh,	purA,	rec ) %>%
  rename(`#Name` = "ST" ) 
  
# metadata
grapeTree_2aux_Borbon_CfxNeg = meta_and_EC_ST__CfxNeg %>% select( ST, Isolate_Type, rMLST_specie ) %>%
  rename( `id` = "ST")


```

# actually save file for grapetree

```{r write_result_file_in_grapeTree_format, eval=F}
# save only when sure

write_tsv( grapeTree_1prof_Borbon_CfxPos, "../RESULT_UPLOAD/grapeTree_1prof_Borbon_CfxPos.tsv" )
write_tsv( grapeTree_2aux_Borbon_CfxPos,  "../RESULT_UPLOAD/grapeTree_2aux_Borbon_CfxPos.tsv")

write_tsv( grapeTree_1prof_Borbon_CfxNeg, "../RESULT_UPLOAD/grapeTree_1prof_Borbon_CfxNeg.tsv" )
write_tsv( grapeTree_2aux_Borbon_CfxNeg,  "../RESULT_UPLOAD/grapeTree_2aux_Borbon_CfxNeg.tsv")

```


# some sanity check
why all node are the same size in grapetree?
see ST frequency distribution 
```{r ST_freq}

meta_and_EC_ST__CfxPos %>% group_by( `ST` ) %>% summarize( ST_count = n())  %>% arrange( desc(ST_count) )
meta_and_EC_ST__CfxNeg %>% group_by( `ST` ) %>% summarize( ST_count = n())  %>% arrange( desc(ST_count) )

```

### below for phyloviz, remove when done with grapetree



# phyloviz seems to expect mlst file to have format like this:
# (mlst --legacy would have number only for the matrix without the gene name repeated)
# or see meta+mlst+phyloviz.Rmd for R code that strip out the characters 

 ST      gki     gtr     murI    mutS    recP    xpt     yqiZ
  1       10      6       6       6       12      13      8
  2       5       4       4       1       2       15      2
  3       5       3       4       1       6       2       1


so geneating a mlst_4_phyloviz.tsv  in the next block


```{r generate_phyloviz_profile_dat_file}
# meta+mlst+phyloviz.Rmd  has some code
# but a bit diff now with this new workflow
# write new code if needed.
```
