---
title: "generate Pathotype table (Borbon Summer 2025)"
date: "2026-06-06"
output:
  html_document: default
  pdf_document: default
---


(code based on meta+mlst+grapetree.Rmd)

Generate pathotype table for Gabrielle's Borbon Summer 2025 dataset
(Jay suggestion bmail 2026.06~05)

dataCache: 
/mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT 

use Abricate_vfdb_summary.csv
don't use mlst.all.tsv ?   but the xlsx used below (for grapetree generation also).

eg desired result table:

Pathotype\source  mom  |  child   |  water
---------  ------------|----------|---------
ExPEC           n (%)  |  n (%)   |  n (%)
ETEC            n (%)  |  n (%)   |  n (%)
EPEC            n (%)  |  n (%)   |  n (%)
EAEC            n (%)  |  n (%)   |  n (%)
DAEC            n (%)  |  n (%)   |  n (%)
EIEC/Shigella   n (%)  |  n (%)   |  n (%)  
STEC/EHEC       n (%)  |  n (%)   |  n (%)      # think STEC would be 0



Does it matter if Ceftriaxone +ve or -ve (CRO+ vs CRO-)? 
Separate table for the two media?
A> Combined for manuscript, separated for self evaluation and highlight interesting points in discussion.  maybe as supplements?



Gabrielle, whatever write up that you already have, if you can send me a copy, that would help me get more context and remember where we left off :D 


inputs:
- meta data for sample source, CRO+ vs CRO- (from "Sample Type" col in the xlsx file, but need to extract CRO+ vs CRO- info from the string),
- abricate vfg (not resfinder) result   for the genes used to define Pathotype.
  https://drive.google.com/file/d/1DD_T15Uz0C_zydz8EeNN8fbX_MPqG5__/view?usp=drive_link 
  /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv    104 rows

~~~~

ExPEC definition.  Use local def from Liseth.  citation?

other Pathotype def: 
Paper by Kelsey J. Jesser, Karen Levy: 
Curr Opin Infect Dis. 2020 October ; 33(5): 372–380. doi:10.1097/QCO.0000000000000665

My annotated copy:
J:\Shared drives\Tin_Ho_gDrv_BL_shareDrv\mph_hw\mph_prj_bioinfo\ExPEC_Ala\ Defining diarrheagenic E. coli.pdf 
https://drive.google.com/drive/u/0/folders/1nKbMtrcIxkGjHMCUsdoQvPtEL1MiLbPS


~~~

gDrive for Borbon_summer_25:
https://drive.google.com/drive/u/0/folders/1eDXqacdExuWbvYvnBdqHIjb1QMkfTAJZ


(
previously 
ran rMLST to determine organism, non E.coli will be fitlered out.  result of rMLST placed in new version of metadata file
GM all samples + species 
https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
[saved to metadata/GM_all_samples+species.tsv]
)



~~~



```{r setup}

library(pacman)
p_load(tidyverse)
p_load( dplyr )
p_load( readr )
p_load( stringr )


```


```{bash}

# path for MLST result
pwd
ls -la /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/metadata

ls -la /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv    
# 104 rows

```


# read inputs

```{r read_inputs}


# metadata info
# still using this metadata file, need to extract CFX+ vs CFX- from "Sample Type" col
metadataXlsx = readxl::read_xlsx( r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\metadata\GM_metadata.xlsx)", sheet="Sample IDs + Susceptibility", na="" )  # enclose windows path with r"(...)"  # strange syntax, but whatever.
metadataXlsx %>% dim()  
#metadataXlsx %>% View()
#metadataXlsx %>% head(5)
metadataXlsx %>% colnames()


#VfdbResultFile = "/mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv"
VfdbResultFile = "c:/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv"          # weasel
VfdbResultTsv  = read_tsv( VfdbResultFile, skip=0, show_col_types=F )  



VfdbResultTsv |> head(5) 
VfdbResultTsv |> colnames()
VfdbResultTsv |> dim()      # 103 rows, 201 cols


# specie info from rMLST 
# from MLST, and some are not e.coli.  determination of e.coli will now be determined by rMLST, 
# with info manually merged into the metadata file 
# GM all samples + species 
# https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
# mostly treat this file for rMLST specie determination using column rMLST_result_full_run
# the `MLST results` was determined when run_mlst.sh let the program determine specie
# (which will be changed by this incarnation/version of this script)
metadataWspecieTsv = read_tsv( "../metadata/GM_all_samples+species.tsv" )
metadataWspecieTsv %>% colnames()  
metadataWspecieTsv %>% dim()       # 104 x 9

```



```{note eval=F}

# run rMLST on fasta that MLST said is not ecoli

```

```{r massage_input_file, eval=T}

# metadata, WgS_ID is saved as number, need it as string to join
metadataDF = metadataXlsx %>% 
  mutate( WGS_ID_as_string = as.character(WGS_ID) ) %>% 
  rename( Sample_Type_ID = Sample_Type ) %>% 
  rename( Isolate_Type = 'Sample Type' )


# GM all samples + species 
# https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
# mostly treat this file for rMLST specie determination using column rMLST_result_full_run
# the `MLST results` was determined when run_mlst.sh let the program determine specie
# (which will be changed by this incarnation/version of this script)
rMLST_specie = metadataWspecieTsv %>% 
  rename( rMLST_specie = rMLST_result_full_run ) %>%
  select( - `rMLST results` )  # this col was manually created G and has some inconsistency, not using


#rMLST_specie |> View()
```



```{r rMLST_filter_nonEC}
# so still need the rMLST result file to determine what row are E.coli and only keep those. ++
# actually, rMLST info in metadataWspecieTsv

# rMLST_specie %>% View()
rMLST_specie %>% dim()   # 104 rows
rMLST_EC    = rMLST_specie %>% filter( rMLST_specie == "Escherichia coli" )

rMLST_EC %>% dim() #  93 rows ie filtered 11 rows (incl 1 NA)
rMLST_EC %>% colnames()
# "Name" col has the isolate ID that match Gabrielle's "user_sample_name" (these 2 cols seems to match, except 126 missing in user_sample_name)

rMLST_nonEC <- rMLST_specie %>% filter( ! grepl( "Escherichia coli", rMLST_specie   ) )  
rMLST_nonEC %>% dim() #  expect 11 !!   (1 is NA)
rMLST_nonEC %>% print( n=12)


```



filenames eg: AI-1608_103_103_assembly.fasta
need to extract       ^^^
as join key to be able to merge metadata info.
use Regular Expression in str_match() to extract file id number


mlst output examples

mlst   AI-1608_206_206_assembly.fasta
AI-1608_206_206_assembly.fasta  ecoli   10      adk(10) fumC(11)        gyrB(4) icd(8)  mdh(8)  purA(8) recA(2)

mlst  --scheme ecoli AI-1608_206_206_assembly.fasta
AI-1608_206_206_assembly.fasta  ecoli   10      adk(10) fumC(11)        gyrB(4) icd(8)  mdh(8)  purA(8) recA(2)


mlst --legacy --scheme ecoli AI-1608_206_206_assembly.fasta
FILE    SCHEME  ST      adk     fumC    gyrB    icd     mdh     purA    recA
AI-1608_206_206_assembly.fasta  ecoli   10      10      11      4       8       8       8       2



--legacy is easiest to parse
grapetree and phyloviz likely just want the number (for matrix)

/mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv


```{r prep_joinKeys}

mlstResultDF_unfiltered = mlstResultTsv %>% mutate(
  isolateKey = str_match( FILE, "AI[\\-]1608[\\_]([:digit:]+)[\\_]([:digit:]+)_assembly\\.fasta"  )[,2] 
) %>%  rename( forced_EC_scheme = SCHEME )   # to be very clear MLST output were all forced to E.coli

```


```{r join_metadata_and_mlst}

meta_and_mlst_DF = full_join(
  x  = metadataDF,
  y  = mlstResultDF_unfiltered, #mlstResultDF,
  by = c( `WGS_ID_as_string` = "isolateKey" )
)


meta_and_mlst_DF %>% head()
meta_and_mlst_DF %>% dim()   # 104... some might not be EC
meta_and_mlst_DF |> View()

#0613 zzz
```

```{r join_rMlst}

# filtered rMLST, right join, so non EC would be removed
# expect 93 rows

meta_and_EC_ST = right_join(
  x  = meta_and_mlst_DF,
  y  = rMLST_EC,
  by = c( "WGS_ID" = "Name")
)


meta_and_EC_ST %>% dim()  # 93 rows, good
meta_and_EC_ST %>% colnames() 
meta_and_EC_ST |>  View()       # __key_DF__ __DW_DF__ 


```

# write the combined mlst + metadata file into a new result file
# this was in the old  meta+mlst+grapetree.Rmd, not needed here
```{r save_result, eval=F}


##old previos version## write_tsv( meta_and_mlst_DF, "../Result/meta_and_mlst.tsv" )


## saving new "v2" result // this is optional, in case she wants to look at it.  __key_DF__ __DW_DF__

# write_tsv( meta_and_EC_ST, "../Result/meta_and_EC_ST.tsv" )  
# avoid saving result to public git repo till paper is published
# so save to temporary location and upload to gDrv: https://drive.google.com/drive/u/0/folders/1v-GvAj3N6EZ4LbPpRg1X9DUBhLPn-wFw
write_tsv( meta_and_EC_ST,  r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\Result\joinedRmdDF.tsv)" ) 

```


# massage VFDB table 
simplify, convert to gene present/absent
based on > 80% score as Match.

```{r getFirstNum, eval=T}

## abricate sometime have match value containing semicolon, eg
##  98.75;99.00
##  87.91;100.00;100.00;100.00;100.00
## 100.00;100.00;100.00;100.00             # usda_summary.csv:GCA_052631835.1_PDT002905087.1_genomic_usda.txt
## need to parse them out, maybe keep the first number, up to the first ;

## ref: Env_EcO157/code/stInfo_Snp1.Rmd
## ref: L348, L354.
## ./guatemala_amr/guat_amr_analy.Rmd:## semicolon shows up when multiple match in a given file.
## ./guatemala_amr/guat_amr_analy.Rmd:# many of the genes col could have semicolon separate numbers, eg


# abricate --sumary produce output where
# many of the genes col could have semicolon separate numbers, eg 98.75;99.00
# taking first ; ignore rest, convert to double.
# This kind of output happens when the file has multiple section where it matched.
# running the abricate --db jgrg db/amnt/sequence  will demo this output really well.
# define helper fn that will be used by across() inside a mutate().
getFirstNum = function( charStr ) {
  firstStr = str_match(  charStr , "([:digit:]+\\.*[:digit:]*)")[,1]
  if_else(
    as.numeric(  firstStr ) >= 80 , 1, 0, missing=0     # 80% as cutoff to Positive(1), else Negative(0)
  )
}
```


# join with abricate VFDB result.
# should do this earlier.

```{r massage_vfdb_result, eval=T}
#// vfdb_sum_df_BAK = vfdb_sum_df
#//usdaDB_sum_df = usdaDB_sum_df_text 



VfdbResultDf = VfdbResultTsv     |> 
  rename( file_name = `#FILE` )  |>
  mutate( across( 4:length( VfdbResultTsv)-1, getFirstNum ))  
#                 ^-- col idx 4 was the final index used by Env_EcO157/code/stInfo_Snp1.Rmd

VfdbResultTsv |> head(5)
VfdbResultDf  |> head(5)
#VfdbResultDf |> View()
VfdbResultTsv |> dim()    # 103x201
VfdbResultDf  |> dim()    # 103x201
  
```

```{r determinePathotype}

pathotypeDf = VfdbResultDf |>
  mutate(
    ExPEC = if_else( ExPEC__papA + ExPEC__papC + ExPEC__sfaS + ExPEC__kpsMII + ExPEC__iutA >= 2, 1, 0, missing=0 ),   # ExPEC definition: at least 2 of the 5 genes
    ETEC  = if_else( ETEC__lt + ETEC__st >= 1, 1, 0, missing=0 ),   # ETEC definition: at least 1 of the 2 genes
    EPEC  = if_else( EPEC__eae + EPEC__bfpA >= 1, 1, 0, missing=0 ),   # EPEC definition: at least 1 of the 2 genes
    EAEC  = if_else( EAEC__aaiC + EAEC__aggR >= 1, 1, 0, missing=0 ),   # EAEC definition: at least 1 of the 2 genes
    DAEC  = if_else( DAEC__daaD >= 1, 1, 0, missing=0 ),   # DAEC definition: at least 1 of the gene
    EIEC_Shigella = if_else( EIEC_Shigella__ipaH >= 1, 1, 0, missing=0 ),   # EIEC/Shigella definition: at least 1 of the gene
    STEC_EHEC = if_else( STEC_EHEC__stx1 + STEC_EHEC__stx2 >= 1, 1, 0, missing=0 )   # STEC/EHEC definition: at least 1 of the stx genes
    )

# while I was looking at the paper, copilot came up with the above.  i was going to use case when.  the gene names are wrong.  the ExPEC logic... did i use it before?
# search email ...
# https://doi.org/10.1038/s41579-020-0416-x  // dont look like local def
# Saba had a UPEC Local and Global... manuscript (not out yet) , it states: 
# study, a different molecular definition by Johnson et al. for the broader ExPEC group (presence
# of ≥2 of papAH and/or papC, sfa/focDE, afa/dra, iutA, kpsMII) was the strongest predictor of
# experimental virulence in a murine model.

```


## split CFX+ vs CFX-  


```{r split_cfx_pos_vs_neg}

# ++ likely need to change to use new DF with VFG info ++
meta_and_EC_ST__CfxPos = meta_and_EC_ST %>%
  filter(   grepl( "Ceftriaxone", Isolate_Type, ignore.case=T ) ) 

meta_and_EC_ST__CfxNeg = meta_and_EC_ST %>%
  filter( ! grepl( "Ceftriaxone", Isolate_Type, ignore.case=T ) ) 


meta_and_EC_ST__CfxPos %>% dim()  # 37
meta_and_EC_ST__CfxNeg %>% dim()  # 56

# meta_and_EC_ST__CfxNeg %>% View()
```



