---
title: "generate Pathotype table (Borbon Summer 2025)"
date: "2026-06-06"
output:
  html_document: default
  pdf_document: default
---


(code based on meta+mlst+grapetree.Rmd)

Generate pathotype table for Gabrielle's Borbon Summer 2025 dataset
(Jay suggestion bmail 2026.06~05)

dataCache: 
/mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT 

use Abricate_vfdb_summary.csv
don't use mlst.all.tsv ?   but the xlsx used below (for grapetree generation also).

eg desired result table:

Pathotype\source  mom  |  child   |  water
---------  ------------|----------|---------
ExPEC           n (%)  |  n (%)   |  n (%)
ETEC            n (%)  |  n (%)   |  n (%)
EPEC            n (%)  |  n (%)   |  n (%)
EAEC            n (%)  |  n (%)   |  n (%)
DAEC            n (%)  |  n (%)   |  n (%)
EIEC/Shigella   n (%)  |  n (%)   |  n (%)  
STEC/EHEC       n (%)  |  n (%)   |  n (%)      # think STEC would be 0



Does it matter if Ceftriaxone +ve or -ve (CRO+ vs CRO-)? 
Separate table for the two media?
A> Combined for manuscript, separated for self evaluation and highlight interesting points in discussion.  maybe as supplements?



Gabrielle, whatever write up that you already have, if you can send me a copy, that would help me get more context and remember where we left off :D 


++ Need to check denominator count.  E.coli?  what rows had been filtered?  ++

inputs:
- meta data for sample source, CRO+ vs CRO- (from "Sample Type" col in the xlsx file, but need to extract CRO+ vs CRO- info from the string),
- abricate vfg (not resfinder) result   for the genes used to define Pathotype.
  https://drive.google.com/file/d/1DD_T15Uz0C_zydz8EeNN8fbX_MPqG5__/view?usp=drive_link 
  /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv    104 rows

~~~~

ExPEC definition.  Use local def from Liseth.  citation?

other Pathotype def: 
Paper by Kelsey J. Jesser, Karen Levy: 
Curr Opin Infect Dis. 2020 October ; 33(5): 372–380. doi:10.1097/QCO.0000000000000665

My annotated copy:
J:\Shared drives\Tin_Ho_gDrv_BL_shareDrv\mph_hw\mph_prj_bioinfo\ExPEC_Ala\ Defining diarrheagenic E. coli.pdf 
https://drive.google.com/drive/u/0/folders/1nKbMtrcIxkGjHMCUsdoQvPtEL1MiLbPS


~~~

gDrive for Borbon_summer_25:
https://drive.google.com/drive/u/0/folders/1eDXqacdExuWbvYvnBdqHIjb1QMkfTAJZ


(
previously 
ran rMLST to determine organism, non E.coli will be fitlered out.  result of rMLST placed in new version of metadata file
GM all samples + species 
https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
[saved to metadata/GM_all_samples+species.tsv]
)



~~~



```{r setup}

library(pacman)
p_load(tidyverse)
p_load( dplyr )
p_load( readr )
p_load( stringr )


```


```{bash}

# path for MLST result
pwd
ls -la /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/metadata

ls -la /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv    
# 104 rows

```


# read inputs

```{r read_inputs}


# metadata info
# still using this metadata file, need to extract CFX+ vs CFX- from "Sample Type" col
#metadataXlsx = readxl::read_xlsx( r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\metadata\GM_metadata.xlsx)", sheet="Sample IDs + Susceptibility", na="", col_types="text" )  # HORRIBLE! col_types="text"  convert WGS_ID to 103.0 !!  wtf?!
metadataXlsx = readxl::read_xlsx( r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\metadata\GM_metadata.xlsx)", sheet="Sample IDs + Susceptibility", na="" )  # enclose windows path with r"(...)"  # strange syntax, but whatever.
metadataXlsx %>% dim()  
#metadataXlsx %>% View()
metadataXlsx %>% head(5)
metadataXlsx %>% colnames()
metadataXlsx %>% str()

# read MLST result file (now use the forced E.coli method)
#                   /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv
mlstResultFile       = "c:/tin/dataCache/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv"
mlstResultTsv   = read_tsv( mlstResultFile, skip=0, show_col_types=F, col_types=cols(.default="c") )        # sample records... isolate info    ## eventually to make to Aux file
# col_types=cols(.default="c") force to character, no need to parse any of the col as number

#mlstResultTsv |> str()

# specie info from rMLST 
# from MLST, and some are not e.coli.  determination of e.coli will now be determined by rMLST, 
# with info manually merged into the metadata file 
# GM all samples + species 
# https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
# mostly treat this file for rMLST specie determination using column rMLST_result_full_run
# the `MLST results` was determined when run_mlst.sh let the program determine specie
# (which will be changed by this incarnation/version of this script)
metadataWspecieTsv = read_tsv( "../metadata/GM_all_samples+species.tsv", skip=0, show_col_types=F, col_types=cols(.default="c") )  # here type forcing is fine, still have 103 without .0 added. phew!
metadataWspecieTsv %>% colnames()  
metadataWspecieTsv %>% dim()       # 104 x 9
metadataWspecieTsv |>  str()
metadataWspecieTsv |>  head()

#VfdbResultFile = "/mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv"
VfdbResultFile = "c:/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv"          # weasel
VfdbResultTsv  = read_tsv( VfdbResultFile, skip=0, show_col_types=F, col_types=cols(.default="c") )           # NUM_FOUND could be a number, but not really computing on it anyway

VfdbResultTsv |> head(5) 
#VfdbResultTsv |> colnames()
VfdbResultTsv |> dim()      # 103 rows, 201 cols


```



```{note eval=F}

# run rMLST on fasta that MLST said is not ecoli

```

```{r massage_input_file, eval=T}

# metadata, WgS_ID is saved as number, need it as string to join
metadataDF = metadataXlsx %>% 
  mutate( WGS_ID_as_string = as.character(WGS_ID) ) %>%   # forced input col type didn't work, so keep this.
  rename( Sample_Type_ID = Sample_Type ) %>% 
  rename( Isolate_Type = 'Sample Type' ) |>
  mutate(
    source_water  = if_else( grepl( "Water",  Isolate_Type, ignore.case=T),  1, 0, missing=0 ),
    source_mother = if_else( grepl( "Mother", Isolate_Type, ignore.case=T),  1, 0, missing=0 ),
    source_child  = if_else( grepl( "Child",  Isolate_Type, ignore.case=T),  1, 0, missing=0 )
  ) |> mutate(
    QA_source_count = ( source_water + source_mother + source_child )  # expect 1 for every row, check if not!
  )


metadataDF |> filter( QA_source_count != 1 )   # they are indeed all 1, pattern match worked.
#metadataDF |> head(5)
#metadataDF |> View()



# GM all samples + species 
# https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
# mostly treat this file for rMLST specie determination using column rMLST_result_full_run
# the `MLST results` was determined when run_mlst.sh let the program determine specie
# (which will be changed by this incarnation/version of this script)
rMLST_specie = metadataWspecieTsv %>% 
  rename( rMLST_specie = rMLST_result_full_run ) %>%
  select( - `rMLST results` )  # this col was manually created G and has some inconsistency, not using


#rMLST_specie |> View()
```



```{r rMLST_filter_nonEC}
# so still need the rMLST result file to determine what row are E.coli and only keep those. 
# actually, rMLST info in metadataWspecieTsv

# rMLST_specie |> str()
# rMLST_specie %>% View()
rMLST_specie %>% dim()   # 104 rows
rMLST_EC    = rMLST_specie %>% filter( rMLST_specie == "Escherichia coli" )

rMLST_EC %>% dim() #  93 rows ie filtered 11 rows (incl 1 NA)
rMLST_EC %>% colnames()
# "Name" col has the isolate ID that match Gabrielle's "user_sample_name" (these 2 cols seems to match, except 126 missing in user_sample_name)

rMLST_nonEC <- rMLST_specie %>% filter( ! grepl( "Escherichia coli", rMLST_specie   ) )  
rMLST_nonEC %>% dim() #  expect 11 !!   (1 is NA)
rMLST_nonEC %>% print( n=12)


```



filenames eg: AI-1608_103_103_assembly.fasta
need to extract       ^^^
as join key to be able to merge metadata info.
use Regular Expression in str_match() to extract file id number


mlst output examples

mlst   AI-1608_206_206_assembly.fasta
AI-1608_206_206_assembly.fasta  ecoli   10      adk(10) fumC(11)        gyrB(4) icd(8)  mdh(8)  purA(8) recA(2)

mlst  --scheme ecoli AI-1608_206_206_assembly.fasta
AI-1608_206_206_assembly.fasta  ecoli   10      adk(10) fumC(11)        gyrB(4) icd(8)  mdh(8)  purA(8) recA(2)


mlst --legacy --scheme ecoli AI-1608_206_206_assembly.fasta
FILE    SCHEME  ST      adk     fumC    gyrB    icd     mdh     purA    recA
AI-1608_206_206_assembly.fasta  ecoli   10      10      11      4       8       8       8       2



--legacy is easiest to parse
grapetree and phyloviz likely just want the number (for matrix)

/mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv


```{r prep_joinKeys}

mlstResultDF_unfiltered = mlstResultTsv %>% mutate(
  isolateKey = str_match( FILE, "AI[\\-]1608[\\_]([:digit:]+)[\\_]([:digit:]+)_assembly\\.fasta"  )[,2] 
) %>%  rename( forced_EC_scheme = SCHEME )   # to be very clear MLST output were all forced to E.coli

```


```{r join_metadata_and_mlst}

#metadataDF |> str()
metadataDF |> dim()  # 104
metadataDF |> head()  # 104
metadataDF |> select( House_ID, Sample_Type_ID, WGS_ID, WGS_ID_as_string ) |> head()

mlstResultDF_unfiltered |> dim() # 103
mlstResultDF_unfiltered |> head()
mlstResultDF_unfiltered |> select( FILE, forced_EC_scheme, isolateKey ) |>  tail()  # id range: 103..206

meta_and_mlst_DF = full_join(
  x  = metadataDF,
  y  = mlstResultDF_unfiltered, #mlstResultDF,
  by = c( `WGS_ID_as_string` = "isolateKey" )
)


meta_and_mlst_DF %>% head()
meta_and_mlst_DF %>% tail()
meta_and_mlst_DF %>% dim()   # 104... some might not be EC
#meta_and_mlst_DF |> View()   # still have House_ID info here   but Sample_Type_ID and WGS_ID , which as character, has .0 added.  weird!!

```

```{r join_rMlst}

# filtered rMLST, right join, so non EC would be removed
# expect 93 rows

# meta_and_mlst_DF |> head()
# meta_and_EC_ST |> View()
meta_and_mlst_DF |> dim() # 104x25   ## for a bit was 207 x 25 # not sure why, debugging shouldn't fix it.  stale cache? restarted R then it is ok.
rMLST_EC |> head()
rMLST_EC |> tail()

rMLST_EC |> dim()  # 93

meta_and_EC_ST = right_join(
  x  = meta_and_mlst_DF,
  y  = rMLST_EC,
  #by = c( "WGS_ID" = "Name")
  by = c( "WGS_ID_as_string" = "Name" )
)


meta_and_EC_ST %>% dim()  # 93 rows, good
meta_and_EC_ST %>% colnames() 
#meta_and_EC_ST |>  str()
meta_and_EC_ST |>  head()          # lost House_ID  info.  reset of R fixed it.
meta_and_EC_ST |>  tail()
# meta_and_EC_ST |>  View()       # __key_DF__ __DW_DF__ 


```

# write the combined mlst + metadata file into a new result file
# this was in the old  meta+mlst+grapetree.Rmd, not needed here
```{r save_result, eval=F}


##old previos version## write_tsv( meta_and_mlst_DF, "../Result/meta_and_mlst.tsv" )


## saving new "v2" result // this is optional, in case she wants to look at it.  __key_DF__ __DW_DF__

# write_tsv( meta_and_EC_ST, "../Result/meta_and_EC_ST.tsv" )  
# avoid saving result to public git repo till paper is published
# so save to temporary location and upload to gDrv: https://drive.google.com/drive/u/0/folders/1v-GvAj3N6EZ4LbPpRg1X9DUBhLPn-wFw
write_tsv( meta_and_EC_ST,  r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\Result\joinedRmdDF.tsv)" ) 

```


# massage VFDB table 
simplify, convert to gene present/absent
based on > 80% score as Match.

```{r getFirstNum, eval=T}

## abricate sometime have match value containing semicolon, eg
##  98.75;99.00
##  87.91;100.00;100.00;100.00;100.00
## 100.00;100.00;100.00;100.00             # usda_summary.csv:GCA_052631835.1_PDT002905087.1_genomic_usda.txt
## need to parse them out, maybe keep the first number, up to the first ;

## ref: Env_EcO157/code/stInfo_Snp1.Rmd
## ref: L348, L354.
## ./guatemala_amr/guat_amr_analy.Rmd:## semicolon shows up when multiple match in a given file.
## ./guatemala_amr/guat_amr_analy.Rmd:# many of the genes col could have semicolon separate numbers, eg


# abricate --sumary produce output where
# many of the genes col could have semicolon separate numbers, eg 98.75;99.00
# taking first ; ignore rest, convert to double.
# This kind of output happens when the file has multiple section where it matched.
# running the abricate --db jgrg db/amnt/sequence  will demo this output really well.
# define helper fn that will be used by across() inside a mutate().
getFirstNum = function( charStr ) {
  firstStr = str_match(  charStr , "([:digit:]+\\.*[:digit:]*)")[,1]
  if_else(
    as.numeric(  firstStr ) >= 80 , 1, 0, missing=0     # 80% as cutoff to Positive(1), else Negative(0)
  )
}
```


# join with abricate VFDB result.
# should do this earlier.

```{r massage_vfdb_result, eval=T}
#// vfdb_sum_df_BAK = vfdb_sum_df
#//usdaDB_sum_df = usdaDB_sum_df_text 



VfdbResultDf = VfdbResultTsv     |> 
  rename( file_name = `#FILE` )  |> mutate(
  filename_sid   = str_match( file_name, "AI-1608_([0-9]+)_([0-9]+)_assembly_Abricate_vfdb.tsv")[,2]   ) |>
  mutate( across( 3:length( VfdbResultTsv), getFirstNum ))  

#  mutate( across( 4:length( VfdbResultTsv), getFirstNum ))  
#                 ^-- col idx and length()-1 are very fiddly.  depends on how read* fn worked?.  


#- VfdbResultDf_test = VfdbResultDf |> mutate( filename_sid   = str_match( file_name, "AI-1608_([0-9]+)_([0-9]+)_assembly_Abricate_vfdb.tsv")[,2]   )
#- VfdbResultDf_test |> View()
# extract sample ID from file_name, so that it can later be joined to meta data WGS_ID   eg 155
# file_name from vfdb result eg: 
#  AI-1608_155_155_assembly_Abricate_vfdb.tsv
  
VfdbResultTsv |> head(5)
VfdbResultDf  |> head(5)
#VfdbResultDf |> View()
VfdbResultTsv |> dim()    # 103x201
VfdbResultDf  |> dim()    # 103x201
VfdbResultDf  |> str()  
VfdbResultDf  |> select( length(VfdbResultDf)-180 : length(VfdbResultDf) ) |> str()
VfdbResultDf  |> select( 190 : length(VfdbResultDf) ) |> str()
VfdbResultDf  |> colnames()  
```

```{r determinePathotype}

pathotype_names = c( "ExPEC", "ETEC", "EPEC", "STEC", "EAEC", "DAEC", "EIEC" )  # "EIEC_Shigella",  "STEC_EHEC"

ETEC_genes = c( "lt", "sta" )
EPEC_genes = c( "eae" )         # this dataset has only eae
STEC_genes = c( "stx" )         # no stx* in vfdb result
EIEC_genes = c( "ipaH" )        # no ipa* in vfdb result
EAEC_genes = c( "aatA", "aggR", "aaiC" )        # OR condition, none exist in vfdb result
#DAEC_genes = c( "afa", "dra", "daa" )           # OR condition, names in VFDB result are more specific
DAEC_genes = c( "afaA", "afaB-I", "afaC-I", "afaD", "afaE-II", "afaF-III", "draA", "draB", "draC", "draD", "draE2", "draP", "daaD", "daaE", "daaF" )

# most gene definition from 
# doi:10.1097/QCO.0000000000000665   // my annotated file: file:///J:/Shared%20drives/Tin_Ho_gDrv_BL_shareDrv/mph_hw/mph_prj_bioinfo/ExPEC_Ala/Defining%20diarrheagenic%20E.%20coli.pdf 
# Table 1 p 16 
# DAEC said: Genes in the afa, dra, or daa adhesin operons; cross-reactivity with EAEC genes  (and that more work needed for better definition).  so just say predicted possible pathotype.
# Table 2 p 17, paper reported afaE-I or afaE-III


ExPEC_genes = c( "papA","papA","papB", "papC","papD","papE","papF","papG","papH" , "sfaC","sfaY", "kpsMII", "iutA" )   # only kpsM, no kpsMII

pathotypeDf = VfdbResultDf |>
  mutate(
    #ExPEC = if_else( papA + papC + sfaC + kpsMII + iutA >= 2, "MoreWorkNeeded", 0, missing=0 ),   # ExPEC definition: at least 2 of the 5 genes
    #ETEC  = if_else( lt + sta >= 1, "lt/ltA check", 0, missing=0 ),   # ETEC definition: at least 1 of the 2 genes   # no lt in this dataset
    #ETEC  = if_else(  if_any(any_of("lt"))  + if_any(any_of("sta")) >= 1, "lt/ltA check", 0, missing=0 ),   # ETEC definition: at least 1 of the 2 genes   # no lt in this dataset
    ExPEC = if_else(  rowSums( pick( any_of( ExPEC_genes)), na.rm=T  ) >= 2, 0.9, 0, missing=-1 ),     # ExPEC definition: at least 2 of 5 genes, and more nuanced... more work needed ++
    ETEC  = if_else(  rowSums( pick( any_of( ETEC_genes )), na.rm=T  ) >= 2, 1, 0, missing=-1 ),   # ETEC definition: at least 1 of the 2 genes   # this seems to work.  no lt or sta in this dataset.
    EPEC  = if_else(  rowSums( pick( any_of( EPEC_genes )), na.rm=T  ) >= 1, 1, 0, missing=-1 ),
    STEC  = if_else(  rowSums( pick( any_of( STEC_genes )), na.rm=T  ) >= 1, 1, 0, missing=-1 ),
    EIEC  = if_else(  rowSums( pick( any_of( EIEC_genes )), na.rm=T  ) >= 1, 1, 0, missing=-1 ),
    EAEC  = if_else(  rowSums( pick( any_of( EAEC_genes )), na.rm=T  ) >= 1, 1, 0, missing=-1 ),   # or condition of 3 genes, thus still just checking for >=1 
    DAEC  = if_else(  rowSums( pick( any_of( DAEC_genes )), na.rm=T  ) >= 1, 1, 0, missing=-1 )    # or condition of 3 genes, thus still just checking for >=1 
    )

#pathotypeDf |> select( file_name, ETEC  )  |> head(10)
#pathotypeDf |> filter( ETEC != "No" ) |> select( file_name, ETEC  )  |> head(10)  # 0 rows
#pathotypeDf |> filter( EPEC != "No" ) |> select( file_name, EPEC  )  |> head(10)  # 1 rows

pathotypeDf |> filter( if_any( any_of( pathotype_names ), ~ .x != 0 )) |> select( file_name, filename_sid, any_of( pathotype_names)  )  |> head(25)  # 21 rows

#pathotypeDf |> View()

# while I was looking at the paper, copilot came up with the above.  i was going to use case when.  the gene names are wrong.  the ExPEC logic... did i use it before?
# search email ...
# https://doi.org/10.1038/s41579-020-0416-x  // dont look like local def
# Saba had a UPEC Local and Global... manuscript (not out yet) , it states: 
# study, a different molecular definition by Johnson et al. for the broader ExPEC group (presence
# of ≥2 of papAH and/or papC, sfa/focDE, afa/dra, iutA, kpsMII) was the strongest predictor of
# experimental virulence in a murine model.

```


# save pathotypeDf   and color cell in spreadsheet to highlight the yes
# include Ceftriaxone status   so can just use spreadsheet to group/filter.  not too many records.



```{r pathotype_meta_join}

pathotype_meta = full_join(
  x  = meta_and_EC_ST,
  y  = pathotypeDf,  
  by = c( `WGS_ID_as_string`  = "filename_sid" )  
  #by = c( "WGS_ID" = "filename_sid" )  
  #by = c( "FILE" = "file_name" )
) |> rename( CRO = CTX_class_binary )

  # |> mutate(
  # CRO = case_when( 
  #  grepl( "Ceftriaxone", Isolate_Type, ignore.case=T ) ~ "Positive",
  #  TRUE ~ "Negative" ) )
  # already have CTX_class_binary

# file_name from vfdb result eg: AI-1608_155_155_assembly_Abricate_vfdb.tsv
# FILE                eg AI-1608_103_103_assembly.fasta
# file_name           eg 
# fasta_used_by_rMLST eg AI-1608_103_103_assembly

# pathotype_meta |>  View()  # ~10 rows have NA, check with Gabrielle which NA are expected, which rows to filter, what to keep.



# sanity check

meta_and_EC_ST |> colnames()

#pathotype_meta |> select( file_name, WGS_ID_as_string, any_of( pathotype_names) ) |> head(25)
pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, any_of( pathotype_names), CRO ) |> head(25)
#pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, forced_EC_scheme, `MLST results`, rMLST_specie, forced_EC_scheme, ST, any_of( pathotype_names), CRO ) |> View()
# pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, any_of( pathotype_names), CRO ) |> View()


pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, forced_EC_scheme, `MLST results`, rMLST_specie, forced_EC_scheme, ST, any_of( pathotype_names), CRO ) |> filter( is.na( House_ID))
# why these 10 rows has NA ? 
# 93 rows, remaining 10 don't have ST, rMLST_specie info, etc.

```


```{r write_summary_pathotype_result, eval=F}

sum_pathotype_result_brief = pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, any_of( pathotype_names), CRO ) 

sum_pathotype_result = pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, forced_EC_scheme, `MLST results`, rMLST_specie, forced_EC_scheme, ST, any_of( pathotype_names), CRO ) 


write_tsv( sum_pathotype_result, "../Result/pathotype.tsv" )

#0614 draft result.  ExPEC need logic update.  rest of pathotype need review with Jay on logic being acceptable. ++



```


gDrive for Borbon data, pathotype result uploaded there
https://drive.google.com/drive/u/0/folders/1eDXqacdExuWbvYvnBdqHIjb1QMkfTAJZ


have more columns from metadata and VFG that can be included as needed

Methods write up:


In-silico prediction of pathotype.  
Abricate is run on the WGS of the isolates to determine the virulence genes each isolate carries.
Genes that predict Pathotypes as described in Jesser and Levy (2020) [ doi:10.1097/QCO.0000000000000665 ] is screened for.
Briefly, 
presence of both _lt_ and _sta_ genes would be predictive of ETEC pathotype (none in this dataset).  
Presence of _eae_ genes would be predictive of EPEC pathotype (1 isolate in this dataset).
Presence of _stx_ genes would be preditive of STEC pathotype (none in this dataset).
Presence of _ipaH_ genes would be preditive of EIEC pathotype (none in this dataset).
Presence of any of _aatA_, _aggR_ or _aaiC_, genes would be preditive of EAEC pathotype (none in this dataset).
Presence of any of _afa_, _dra_ or _daa_ family of genes would be preditive of DAEC pathotype (~6 this dataset).

For ExPEC, we used a local definition (paper for reference) [More work TBD]

Please note that the pathotype is predicted from computational screening of DNA sequences, and, as stated by Jesser and Levy, some of these molecular based pathotype definition can use of further refinements.  Thus, please use these predicted pathotypes with caution.



(Figure caption, or even table headings, should emphasize it is predicted pathotype.)




```{r pathotype_prevalence, eval=T}
# Need a table as agreed, which have counts (and %)
# 

# also, check that 93 rows with data is the correct count ++


### make a smaller 2D_table first
Patho_Source_DF = pathotype_meta |> select( House_ID, WGS_ID_as_string, source_mother, source_child, source_water, rMLST_specie, any_of( pathotype_names), CRO ) 
#Patho_Source_DF |> View()

Patho_Source_DF_long = Patho_Source_DF |> pivot_longer( pathotype_names, names_to="Pathotype", values_to="Prevalence" )
# Patho_Source_DF_long |> View()

PathoPrevChild  = Patho_Source_DF_long |> filter( source_child  == 1 ) |> group_by( Pathotype ) |> summarize( Child_Prev = sum(Prevalence))
PathoPrevMother = Patho_Source_DF_long |> filter( source_mother == 1 ) |> group_by( Pathotype ) |> summarize( Mother_Prev = sum(Prevalence))
PathoPrevWater  = Patho_Source_DF_long |> filter( source_water  == 1 ) |> group_by( Pathotype ) |> summarize( Water_Prev = sum(Prevalence))
# 16 MaybeExPEC, 2 from water , 7 each mother and child, add up right
# above split in 3, if concat, should be the right result

# PathoPrevChild

PathoPrevTemp = left_join(
  x  = PathoPrevMother, 
  y  = PathoPrevChild
)

PathoPrevAll = left_join(
  x  = PathoPrevTemp, 
  y  = PathoPrevWater
)

PathoPrevAll    #0614   this table decent, pasted to gSheet for discussion
# https://docs.google.com/spreadsheets/d/1sxljIo1sj4x9fDSP2lfo0vXg6DMgh8vkq833sdX3-UE/edit?gid=1178728147#gid=1178728147


# ++ OH... Need to check denominator count.  E.coli?  what rows had been filtered?  ++

# if desire to find prevalence count by CRO +ve vs -ve, 
# do a Patho_Source_DF_long |> filter( CRO ... ) 
# and recreate the table.

```

```{r double_pivot_longer_create_problem__dont_use, eval=F}
Patho_Source_DF_long_src = Patho_Source_DF_long |> pivot_longer( c(`source_mother`, `source_child`, `source_water` ), names_to="Source", values_to="Source_count" ) # dont like name...
#^^ this second pivot may be messing with values.  creating extra counts... thus the problem.
#^^ instead of creating 3 cols for source_mother, source_child, source_water, create a single col with 3 names, then don't have to pivot twice.  #0614



Patho_Source_DF_long_src  |> group_by( Pathotype, Source ) |> summarize( PathoPrev = sum(Prevalence))
#^^ result is wrong this way!!   sum() didn't split by the 2nd group_by... 

# deleting rest of pathotype prevalence count code that didn't produce correct result, 
# cuz the 2nd pivot_long induced extra counts and corrupt the process
# if needed, see git log e7b7ca2  for the old code
```




```{r litmus_test}

print( THE_END )   # expect error here :D

```


