---
title: "analysis for Shrimp Ecuador 2023 data"
date: "2024-07-13"
output:
  html_document: default
  pdf_document: default
---



Tilden started a maper, MDPI reviwer has request...
data submission to GenBank, Niko found missing fasta for the "I" sample.

trying to reproduce some of the work.
by start "feeling" the result from mlst, etc.


```{r setup}

library( pacman )
p_load( tidyverse )
p_load( dplyr )
p_load( readr )
p_load( stringr )


# fn alias 
`%nin%` = Negate(`%in%`)

# open all input files here in the setup block
# so know early on if have all necessary


####
#### read input, which is output from Abricate
####



# DATADIR="/global/scratch/users/tin/fc_graham/ecuador_2023_TJ/Sequences/Tilden/ALL/assembled-sequences_par3"  # sav
DATADIR="result/"


####
#### MLST SeqType info
####
mlst_tsv_file = sprintf( "%s/%s", DATADIR, "mlst.all.tsv") 
mlst_tsv = read_tsv( mlst_tsv_file )


####
#### resistance gene were focus of paper
####
resfind_sum_file = sprintf( "%s/%s", DATADIR, "Abricate_resfinder_summary.csv") 
resfind_sum_tmp_tsv = read_tsv( resfind_sum_file, col_types = list( .default=col_character() ) )
resfind_sum_tsv = type_convert( resfind_sum_tmp_tsv )  # here, expect only NUM_FOUND is converted as double.


resfind_sum_pct_df = resfind_sum_tsv %>%
  rename( file_name = '#FILE' ,
          RESIST_NUM_FOUND = "NUM_FOUND"
          ) 


####
##### not sure if VFDB was needed.  
####
vfdb_sum_file = sprintf( "%s/%s", DATADIR, "Abricate_vfdb_summary.csv") 
vfdb_sum_tmp_tsv = read_tsv( vfdb_sum_file, col_types = list( .default=col_character() ) )
vfdb_sum_tsv = type_convert( vfdb_sum_tmp_tsv )  # here, expect only NUM_FOUND is converted as double.


# ecoli_vf said to be vfdb + gene from literatur, so more extensive https://github.com/phac-nml/ecoli_vf
# trying it hoping as plug-in replacement to vfdb_sum_tsv
#// ecvf_sum_file = sprintf( "%s/%s", DATADIR, "ecoli_vf_summary.csv" )
#// ecvf_sum_tsv = read_tsv( ecvf_sum_file )

# Choose: VFDB vs ECVF for source of vfg
# Using VFDB

abricate_vfdb_sum_tsv = vfdb_sum_tsv    # this results 155 rows # hmmm... 124 rows in 2024.01 on wombat
#abricate_vfdb_sum_tsv = ecvf_sum_tsv   # code support this now, but result list is ~441 rows



# rest of code will be unchanged to use vfdb_sum_df
# 2024.0315, vfdb_sum_df get the jgrg result joined into it 

vfdb_sum_pct_df = abricate_vfdb_sum_tsv %>%
  rename( file_name = '#FILE' ,
          VFDB_NUM_FOUND = "NUM_FOUND"
          ) 


# View(vfdb_sum_df)

##############################################################################

# reading in files from Tilden
# shared folder:
# https://drive.google.com/drive/folders/1kk-56BYRkYOXJtascCi6LEkVxE-mO0aW
#
# files:
# https://docs.google.com/spreadsheets/d/18qKB7kIbr1adVEm0Ws56eiJPc1ZCX6T55yMw-h10PT8/edit?gid=646212993#gid=646212993
# https://docs.google.com/spreadsheets/d/15P318xrNscGkKK-ls2IUuM3HVyXZy7BhNdxrvVPiYZw/edit?gid=0#gid=0

# ref: https://docs.google.com/document/d/1mP5gzfIKnpeE80osPNdE_b_Gev6pQ6_elGxjC7httpM/edit 

CACHEDIR="CACHE"  # tmp subfolder, files not checked into git repo


SummaryTableXlsx     = sprintf( "%s/Summary Table .xlsx", CACHEDIR ) 
SummaryTableTsv      = sprintf( "%s/Summary Table  - Sheet1.tsv", CACHEDIR ) 
#SummaryTable     = readxl::read_xlsx( SummaryTableXlsx )
SummaryTable     = read_tsv( SummaryTableTsv  )  # edited: removed row 1 filename header line
# 36 rows, so not full dataset of paper

ResMlstCombTableXlsx = sprintf( "%s/Res & MLST Combined.xlsx", CACHEDIR ) 
ResMlstCombTableTsv  = sprintf( "%s/Res & MLST Combined - Abricate_resfinder_summary.tsv", CACHEDIR ) 
#ResMlstCombTable = readxl::read_xlsx( ResMlstCombTableXlsx ) %>%
ResMlstCombTable = read_tsv( ResMlstCombTableTsv ) %>%    # edited, remove tailing comments, empty rows
  rename( file_name = '#FILE' )
# 44 rows, likely full dataset of paper.  Sample "I" is here, found to have ST 3580.  0 ARG.   at least filename to sample type, farm type is correct?   ++CHEcK++

```
# create join key for  mlst  resfinder result
likely just the first 1-2 letter of the filename should suffice
eg for mlst:   (has 55 rows, sample "I" is missing)
AA_CKDN230030154-1A_HGKHYDSX7_L2.fasta
 J_CKDN230030143-1A_HGKHYDSX7_L2.fasta
2^ ^^^^3^^^^^^^^ 4^ ^5^^^^^^^ ^6 

eg for resfinder:  (has 56 rows, sample "I" is present)
I_CKDN230030142-1A_HGKHYDSX7_L2_Abricate_resfinder.tsv
2 ^^^^3^^^^^^^^ 4^ ^5^^^^^^^ ^6 7^^^^^^^^^^^^^^^^^




```{r define_join_key}


resfind_sum_pct_df = resfind_sum_pct_df %>%
  #                                                2                 3                    4                  5                      6              7=DB specific
  mutate( sample_name     = str_match( file_name, "([:alpha:]+)[\\-_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX7[\\-_]L([:alnum:]+)\\_(Abricate_resfinder).tsv"  )[,2] )%>%
  mutate( Hnum            = str_match( file_name, "([:alpha:]+)[\\-_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX7[\\-_]L([:alnum:]+)\\_(Abricate_resfinder).tsv"  )[,5] )%>%
  mutate( Lnum            = str_match( file_name, "([:alpha:]+)[\\-_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX7[\\-_]L([:alnum:]+)\\_(Abricate_resfinder).tsv"  )[,6] )%>%
  mutate( key    = str_c( sample_name, "_H", Hnum, "_L", Lnum,  sep="" ) )

# AB is not unique, ditt for AD, maybe others.


mlst_tsv_df = mlst_tsv %>%
  #                                                2                 3                    4                  5                      6              
  mutate( sample_name     = str_match( Filename, "([:alpha:]+)[\\-_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX7[\\-_]L([:alnum:]+).fasta"  )[,2] )%>%
  mutate( Hnum            = str_match( Filename, "([:alpha:]+)[\\-_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX7[\\-_]L([:alnum:]+).fasta"  )[,5] )%>%
  mutate( Lnum            = str_match( Filename, "([:alpha:]+)[\\-_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX7[\\-_]L([:alnum:]+).fasta"  )[,6] )%>%
  mutate( key    = str_c( sample_name, "_H", Hnum, "_L", Lnum,  sep="" ) )
  
  

```

# join mlst with resfinder result

```{r join}

mlst_resfind_df = full_join(
  x  = mlst_tsv_df, 
  y  = resfind_sum_pct_df,
  by = "key"
  #by = c( 'Filename' = "file_name")  # nope, these filenames are not matching!  need to define a key for them.
)

#View( mlst_resfind_df )

```


# condensed tmp result, write result, this is temporarily, so can look at it using spreadsheet, for discussion

noded filenames with dup sample id
AB
AD
AH
AS
AY
BB - 6 rows... dups?
I has NA cuz the Hnum is from table w/o the ST info


```{r write_tsv}

condense_mlst_resfind_df = mlst_resfind_df %>%  
  select( sample_name.y, Hnum.x, Lnum.x, key, Filename, PubMLST_scheme_name, SeqType, RESIST_NUM_FOUND  ) 

write_tsv( condense_mlst_resfind_df, "result/condense_mlst_restfind_df.tsv")
write_tsv(          mlst_resfind_df, "result/mlst_restfind_df.tsv")

#View( condense_mlst_resfind_df )

```

########################################################

# check Tilden's result, hunting for sample "I" 

filename eg for resfinder result
I_CKDN230030142-1A_HGKHYDSX7_L2_Abricate_resfinder.tsv
2 ^^^^3^^^^^^^^ 4^ ^5^^^^^^^ ^6 7^^^^^^^^^^^^^^^^^


```{r TildenCreateJoinKey}


#SummaryTable     = readxl::read_xlsx( SummaryTableFile )
#ResMlstCombTable = readxl::read_xlsx( ResMlstCombTableFile )

#View( SummaryTable )       # use Sample ID:    ae  j 
#View( ResMlstCombTable )   # extract from Filename AE_CKDN230030158-1A_HGKHYDSX7_L2_Abricate_resfinder.tsv


#  define_join_key
ResMlstCombTable = ResMlstCombTable %>%
  #                                                2                 3                    4                  5                      6              7=DB specific
  mutate( sample_name     = str_match( file_name, "([:alpha:]+)[\\-_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX7[\\-_]L([:alnum:]+)\\_(Abricate_resfinder).tsv"  )[,2] )%>%
  #mutate( Hnum            = str_match( file_name, "([:alpha:]+)[\\-_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX7[\\-_]L([:alnum:]+)\\_(Abricate_resfinder).tsv"  )[,5] )%>%
  #mutate( Lnum            = str_match( file_name, "([:alpha:]+)[\\-_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX7[\\-_]L([:alnum:]+)\\_(Abricate_resfinder).tsv"  )[,6] )%>%
  mutate( key    = str_to_lower( sample_name ) ) 

# AB is not unique, ditt for AD, maybe others.


JoinedSumTab = full_join(
  x  = SummaryTable,
  y  = ResMlstCombTable,
  by = c( `Sample ID:` = "key" )
)

View( JoinedSumTab )

write_tsv(          JoinedSumTab, "CACHE/JoinedSumTab.tsv")
# may want a condensed version , strip out the genes cols... 

```

# create new plot 
Reviewer #4 disliked the pie chart of Figure 4.  recommended stacked bar graph instead.
However, Tilden has 34 samples from water, 10 from shrimp.
Not sure which samples were selected.
The joined table above have ~55 rows.



# Notes about sequence files

most fasta files are about 480 - 550 MB.  But a few are 1210 MB!  ls -lS (it sorted by size, largest first)

-rw-rw-r-- 1 tin users 12115595 Nov 15  2023 AK_CKDN230030164-1A_HGKHYDSX7_L2.fasta
-rw-rw-r-- 1 tin users 10740528 Nov 15  2023 AU_CKDN230030174-1A_HGKHYDSX7_L2.fasta
-rw-rw-r-- 1 tin users  9661854 Nov 15  2023 BA_CKDN230030180-1A_HGKHYDSX7_L2.fasta
-rw-rw-r-- 1 tin users  9033176 Nov 15  2023 S_CKDN230030149-1A_HGKHYDSX7_L2.fasta
-rw-rw-r-- 1 tin users  7035141 Nov 15  2023 E_CKDN230030138-1A_HGKHYDSX7_L2.fasta
-rw-rw-r-- 1 tin users  6635702 Nov 15  2023 AE_CKDN230030158-1A_HGKHYDSX7_L2.fasta
-rw-rw-r-- 1 tin users  6448173 Nov 15  2023 AV_CKDN230030175-1A_HGKHYDSX7_L2.fasta
-rw-rw-r-- 1 tin users  5913388 Nov 15  2023 AZ_CKDN230030179-1A_HGKHYDSX7_L2.fasta
-rw-rw-r-- 1 tin users  5693370 Nov 15  2023 F_CKDN230030139-1A_HGKHYDSX7_L2.fasta
...
-rw-rw-r-- 1 tin users  4505363 Nov 15  2023 Z_CKDN230030153-1A_HGKHYDSX7_L2.fasta





