---
title: "analysis for Shrimp Ecuador 2023 data"
date: "2024-07-13"
output:
  html_document: default
  pdf_document: default
---



Tilden started a maper, MDPI reviwer has request...
data submission to GenBank, Niko found missing fasta for the "I" sample.

trying to reproduce some of the work.
by start "feeling" the result from mlst, etc.


```{r setup}

library( pacman )
p_load( tidyverse )
p_load( dplyr )
p_load( readr )
p_load( stringr )


# fn alias 
`%nin%` = Negate(`%in%`)

# open all input files here in the setup block
# so know early on if have all necessary


####
#### read input, which is output from Abricate
####



# DATADIR="/global/scratch/users/tin/fc_graham/ecuador_2023_TJ/Sequences/Tilden/ALL/assembled-sequences_par3"  # sav
DATADIR="result/"


####
#### MLST SeqType info
####
mlst_tsv_file = sprintf( "%s/%s", DATADIR, "mlst.all.tsv") 
mlst_tsv = read_tsv( mlst_tsv_file )


####
#### resistance gene were focus of paper
####
resfind_sum_file = sprintf( "%s/%s", DATADIR, "Abricate_resfinder_summary.csv") 
resfind_sum_tmp_tsv = read_tsv( resfind_sum_file, col_types = list( .default=col_character() ) )
resfind_sum_tsv = type_convert( resfind_sum_tmp_tsv )  # here, expect only NUM_FOUND is converted as double.


resfind_sum_pct_df = resfind_sum_tsv %>%
  rename( file_name = '#FILE' ,
          RESIST_NUM_FOUND = "NUM_FOUND"
          ) 


####
##### not sure if VFDB was needed.  
####
vfdb_sum_file = sprintf( "%s/%s", DATADIR, "Abricate_vfdb_summary.csv") 
vfdb_sum_tmp_tsv = read_tsv( vfdb_sum_file, col_types = list( .default=col_character() ) )
vfdb_sum_tsv = type_convert( vfdb_sum_tmp_tsv )  # here, expect only NUM_FOUND is converted as double.


# ecoli_vf said to be vfdb + gene from literatur, so more extensive https://github.com/phac-nml/ecoli_vf
# trying it hoping as plug-in replacement to vfdb_sum_tsv
#// ecvf_sum_file = sprintf( "%s/%s", DATADIR, "ecoli_vf_summary.csv" )
#// ecvf_sum_tsv = read_tsv( ecvf_sum_file )

# Choose: VFDB vs ECVF for source of vfg
# Using VFDB

abricate_vfdb_sum_tsv = vfdb_sum_tsv    # this results 155 rows # hmmm... 124 rows in 2024.01 on wombat
#abricate_vfdb_sum_tsv = ecvf_sum_tsv   # code support this now, but result list is ~441 rows



# rest of code will be unchanged to use vfdb_sum_df
# 2024.0315, vfdb_sum_df get the jgrg result joined into it 

vfdb_sum_pct_df = abricate_vfdb_sum_tsv %>%
  rename( file_name = '#FILE' ,
          VFDB_NUM_FOUND = "NUM_FOUND"
          ) 


# View(vfdb_sum_df)





```
# create join key for  mlst  resfinder result
likely just the first 1-2 letter of the filename should suffice
eg for mlst:   (has 55 rows, sample "I" is missing)
AA_CKDN230030154-1A_HGKHYDSX7_L2.fasta
 J_CKDN230030143-1A_HGKHYDSX7_L2.fasta
2^ ^^^^3^^^^^^^^ 4^ ^5^^^^^^^ ^6 

eg for resfinder:  (has 56 rows, sample "I" is present)
I_CKDN230030142-1A_HGKHYDSX7_L2_Abricate_resfinder.tsv
2 ^^^^3^^^^^^^^ 4^ ^5^^^^^^^ ^6 7^^^^^^^^^^^^^^^^^

```{r define_join_key}


resfind_sum_pct_df = resfind_sum_pct_df %>%
  #                                                2                 3                    4                  5                      6              7=DB specific
  mutate( sample_name     = str_match( file_name, "([:alpha:]+)[\\-_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX7[\\-_]L([:alnum:]+)\\_(Abricate_resfinder).tsv"  )[,2] )%>%
  mutate( Hnum            = str_match( file_name, "([:alpha:]+)[\\-_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX7[\\-_]L([:alnum:]+)\\_(Abricate_resfinder).tsv"  )[,5] )%>%
  mutate( Lnum            = str_match( file_name, "([:alpha:]+)[\\-_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX7[\\-_]L([:alnum:]+)\\_(Abricate_resfinder).tsv"  )[,6] )%>%
  mutate( key    = str_c( sample_name, "_H", Hnum, "_L", Lnum,  sep="" ) )

# AB is not unique, ditt for AD, maybe others.


mlst_tsv_df = mlst_tsv %>%
  #                                                2                 3                    4                  5                      6              
  mutate( sample_name     = str_match( Filename, "([:alpha:]+)[\\-_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX7[\\-_]L([:alnum:]+).fasta"  )[,2] )%>%
  mutate( Hnum            = str_match( Filename, "([:alpha:]+)[\\-_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX7[\\-_]L([:alnum:]+).fasta"  )[,5] )%>%
  mutate( Lnum            = str_match( Filename, "([:alpha:]+)[\\-_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX7[\\-_]L([:alnum:]+).fasta"  )[,6] )%>%
  mutate( key    = str_c( sample_name, "_H", Hnum, "_L", Lnum,  sep="" ) )
  
  

```

# join mlst with resfinder result

```{r join}

mlst_resfind_df = full_join(
  x  = mlst_tsv_df, 
  y  = resfind_sum_pct_df,
  by = "key"
  #by = c( 'Filename' = "file_name")  # nope, these filenames are not matching!  need to define a key for them.
)

```


# tmp test... bash works!

```{bash block 1, eval=F}
# can run bash inside Rmd, even on win11 (well, have wsl)
hostname
pwd
ls -latr

echo ""
cd /mnt/c/tin
pwd
ls -latr
```


