
---
title: "TJ Ecuador Pintag 2023 seq submission prep to GenBank"
date: "2024-06-09"
output:
  html_document: default
  pdf_document: default
---

massage data to aid sequence submission to GenBank.
eg, pair fasta file with source (water upstream, water downstream, lettuce).

also
geneate unique sample_name using jgrg_ as prefix + fasta filename
isolate_name_alias (eg S10, the first 2-3 chars, before first '_' )

```{r setup}

library( pacman )
p_load( tidyverse )
p_load( dplyr )
p_load( readr )
p_load( stringr )

#xx p_load(collapse)  # not end up using, for replace_NA(vfdb_sum_df)


# fn alias 
`%nin%` = Negate(`%in%`)

# open all input files here in the setup block
# so know early on if have all necessary


####
#### read input about duplicate sequences
####

isolate_list_csv = read_csv( "Sequences_TjEcuador2023.csv" )


#View( isolate_list_csv )
```

# input file content example
isolate
S10_CKDN230030191-1A_HGKHYDSX7_L2.fasta/assembly.fasta
S2_CKDN230030185-1A_HGKHYDSX7_L2.fasta/assembly.fasta
^^ nickname                           ^^^^^^^^^^^^^^^-- strip
   all start with S


```{r genBankPrep}

str( isolate_list_csv )

isolate_list = isolate_list_csv 
#%>% select(`_Isolate`) %>%
#  rename( isolate = `_Isolate` )

#str( isolate_list )

dat4genBankSubmission = isolate_list %>%
  #                                       2                 3                    4                  5                              6           7                          
  mutate( Snum     = str_match( isolate, "(S[:digit:]+)[\\_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX[:digit:][\\-_]L([:alnum:]+)(.fasta)/assembly.fasta" )[,2] ) %>% 
  mutate( Hnum     = str_match( isolate, "(S[:digit:]+)[\\_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX[:digit:][\\-_]L([:alnum:]+)(.fasta)/assembly.fasta" )[,5] ) %>% 
  mutate( Lnum     = str_match( isolate, "(S[:digit:]+)[\\_](CKDN[:digit:]+)[\\-]([:alnum:]+)[\\-_]H([:alnum:]+)DSX[:digit:][\\-_]L([:alnum:]+)(.fasta)/assembly.fasta" )[,6] ) %>% 
  mutate( filename = str_match( isolate, "(S[:digit:]+[\\_]CKDN[:digit:]+[\\-][:alnum:]+[\\-_]H[:alnum:]+DSX[:digit:][\\-_]L[:alnum:]+.fasta)/assembly.fasta" )[,2] ) %>%
  #mutate( isolate_name_alias = str_c( Snum, "_", Hnum, "_L", Lnum,  sep="" ) )
  mutate( isolate_name_alias = str_c( Snum, "_L", Lnum,  sep="" ) )

head( dat4genBankSubmission )

```
```{r writeResult2file}



# table4genBankSubmission = dat4genBankSubmission %>% select( Snum )  # debug, sanity check, S number along is not unique.  but with L number then they will become unique
table4genBankSubmission = dat4genBankSubmission %>% select( filename, isolate_name_alias )

write_tsv( table4genBankSubmission, "result/table4genBankSubmission.tsv")


```