---
title: "generate Pathotype table (Irrigation Exposure 2026)"
date: "2026-08-17"
output:
  html_document: default
  pdf_document: default
---



Generate pathotype table for Irrigation Exposure 2026 dataset
(Chris LeBoa)

dataCache: 
C:\tin\dataCache\LeBoa\
/mnt/c/tin/dataCache/LeBoa/CLeBoa Farm Irrigation Water NCBI submission.xlsx


use Abricate_vfdb_summary.csv
don't use mlst.all.tsv ?   but the xlsx used below (for grapetree generation also).

eg desired result table:

Pathotype\source  mom  |  child   |  water
---------  ------------|----------|---------
ExPEC           n (%)  |  n (%)   |  n (%)
ETEC            n (%)  |  n (%)   |  n (%)
EPEC            n (%)  |  n (%)   |  n (%)
EAEC            n (%)  |  n (%)   |  n (%)
DAEC            n (%)  |  n (%)   |  n (%)
EIEC/Shigella   n (%)  |  n (%)   |  n (%)  
STEC/EHEC       n (%)  |  n (%)   |  n (%)      




keyDF: 

- meta_and_mlst_DF
- ST (from mlst)
- pathotype (DAEC, etc)  ++ need an alt DF where patotype are in single column as string
- phylogroup (TBD)





Some/all of the comments here is/was for Gabrielle's Borbon ::  ++

        code up to ~ L616 should be good for irr_exp26 data.  
        double check pathotype checked correctly would be nice.  
        don't have full ExPEC/UPEC def yet
        below Line 616 is code originally for Gabrielle M Borbon data, not edited/adjusted for this project yet.  but that chunk most likely not needed ++ 
        ++ double check gene names (in column names) eg is stx and not stx1 or stx1a , I think i checked column name when writting the check, which is why didn't use grepl, but it might have been for the previous project, double check for irr_exp26++  (search for STEC_genes, EPEC_genes, near line 414)


Does it matter if Ceftriaxone +ve or -ve (CRO+ vs CRO-)? 
Separate table for the two media?
A> Combined for manuscript, separated for self evaluation and highlight interesting points in discussion.  maybe as supplements?



++ Need to check denominator count.  E.coli?  what rows had been filtered?  ++

inputs:
- meta data for sample source, CRO+ vs CRO- (from "Sample Type" col in the xlsx file, but need to extract CRO+ vs CRO- info from the string),
- abricate vfg (not resfinder) result   for the genes used to define Pathotype.
  https://drive.google.com/file/d/1DD_T15Uz0C_zydz8EeNN8fbX_MPqG5__/view?usp=drive_link 
  /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv    104 rows

~~~~

ExPEC definition.  Use local def from Liseth.  citation?

other Pathotype def: 
Paper by Kelsey J. Jesser, Karen Levy: 
Curr Opin Infect Dis. 2020 October ; 33(5): 372–380. doi:10.1097/QCO.0000000000000665

My annotated copy:
J:\Shared drives\Tin_Ho_gDrv_BL_shareDrv\mph_hw\mph_prj_bioinfo\ExPEC_Ala\ Defining diarrheagenic E. coli.pdf 
https://drive.google.com/drive/u/0/folders/1nKbMtrcIxkGjHMCUsdoQvPtEL1MiLbPS


~~~

gDrive for 

(
previously 
ran rMLST to determine organism, non E.coli will be fitlered out.  result of rMLST placed in new version of metadata file
GM all samples + species 
https://docs.google.com/spreadsheets/d/1DOw-SbNJgH24SSdvmwLZp38c2hqyS_p9kYmgHkt8RBM/edit?gid=1902543731#gid=1902543731 
[saved to metadata/GM_all_samples+species.tsv]
)



~~~



```{r setup}

library(pacman)
p_load(tidyverse)
p_load( dplyr )
p_load( readr )
p_load( stringr )

ZZZZ_rstudio_instance="pathotype_LeBoa"
# see Values section, buttom of list unless have lots of Fn defined
# mac allow multiple Rstudio instance, and checking variable maybe good way to tell which is which.
# being able to set window title would be even better.

```


```{bash}

# path for MLST result
pwd
#ls -la /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/metadata
ls -la "/mnt/c/tin/dataCache/LeBoa/CLeBoa Farm Irrigation Water NCBI submission.xlsx"

ls -la ../result

#ls -la /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv    
# 104 rows

```


```{r input_filenames_def}

#// metaXlsx1_FileName = r"(C:\tin\dataCache\LeBoa\CLeBoa Farm Irrigation Water NCBI submission.xlsx)"   # maybe akin to python u"..."  for unicode string; r for raw string in R?
#// metaXlsx2_FileName = r"(C:\tin\dataCache\LeBoa\CLeBoa Farm Irrigation Water NCBI submission.xlsx)"  # same filename for now, different sheet , may eventually export to tsx and just read that.

# metadata, use this rewritten tsv going forward
meta_FileName = "../metadata/source+filename.tsv"


# is ST# needed here?
#                   /mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv
#mlstResultFile       = "c:/tin/dataCache/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv"
mlstResultFile       = "../result/mlst.all.forceEC.raw.tsv"

#VfdbResultFile = "/mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv"
#VfdbResultFile = "c:/tin/dataCache/Borbon_summer_25_Gabrielle/Abricate_OUT/Abricate_vfdb_summary.csv"          # weasel
VfdbResultFile = "../result/Abricate_vfdb_summary.csv"   # weasel

phylogroupResultFile = "../result/ezclermont.results.tsv"   # weasel, bigmac2


treeFile = "../result/RAxML_bestTree.ECirr_raxml.tre" 

```

# read inputs - metadata

```{r read_input_metadata}


# metadata info

# C:\tin\dataCache\LeBoa\
# /mnt/c/tin/dataCache/LeBoa/CLeBoa Farm Irrigation Water NCBI submission.xlsx


#metadataXlsx = readxl::read_xlsx( r"(C:\tin\dataCache\Borbon_summer_25_Gabrielle\metadata\GM_metadata.xlsx)", sheet="Sample IDs + Susceptibility", na="", col_types="text" )  # HORRIBLE! col_types="text"  convert WGS_ID to 103.0 !!  wtf?!
#metadataXlsx = readxl::read_xlsx( r"(C:\tin\dataCache\LeBoa\CLeBoa Farm Irrigation Water NCBI submission.xlsx)", sheet="OneHealthEnteric.1.0", na="", skip=18 )  # enclose windows path with r"(...)"  # strange syntax, but whatever.

#metadataNfo = readxl::read_xlsx( metaXlsx1_FileName , sheet="OneHealthEnteric.1.0", na="", skip=18 )  
#metadataNfo = readxl::read_xlsx( metaXlsx2_FileName , sheet="filenames", na="", skip=0 )  

metadataNfo = read_tsv( meta_FileName, na="", skip=0 )  

# file 2 has  sample_name, sample_title & fasta_filename, the fasta_filename isn't avail in the xlsx for GenBank... I think eventually it will be needed ... 


metadataNfo %>% dim()       # 74*12
#metadataXlsx %>% View()
#metadataNfo %>% head(5)
#metadataNfo %>% colnames()
metadataNfo %>% str()



```




# Convert metadta from xlsx to tsv - ONE time run only.  
```{r oneTimeWriteOutMetadataDf2tsv, eval=F}
## write metadata DF to tsv - ONE time run only to convert xlsx to tsv
## write_tsv( metadataXlsx, "../metadata/source+filename.tsv" )
```



```{r read_input_abricate}


# read MLST result file (now use the forced E.coli method)
mlstResultTsv   = read_tsv( mlstResultFile, skip=0, show_col_types=F, col_types=cols(.default="c") )        # sample records... isolate info    ## eventually to make to Aux file
# col_types=cols(.default="c") force to character, no need to parse any of the col as number


mlstResultTsv |> dim()  # 72*10
mlstResultTsv |> head()
#mlstResultTsv |> str()


VfdbResultTsv  = read_tsv( VfdbResultFile, skip=0, show_col_types=F, col_types=cols(.default="c") )           # NUM_FOUND could be a number, but not really computing on it anyway

VfdbResultTsv |> head(5) 
#VfdbResultTsv |> colnames()
VfdbResultTsv |> dim()      # 73 rows, 176 cols

```
```{r read_input_pathotype}

# phylogroupResultFile = "../result/ezclermont.results.tsv"   # weasel, bigmac2

# file has no headername
phylogroupResultTsv = read_tsv( phylogroupResultFile, col_names = c( "seq_filename", "phylogroup") )

phylogroupResultTsv |> str()

phylogroupResultDf = phylogroupResultTsv |> mutate(
  seqCo_sid      = str_match( seq_filename, "AI-5395_([0-9]+)_([0-9]+)_assembly")[,2] ,   
  filename_sid   = str_match( seq_filename, "AI-5395_([0-9]+)_([0-9]+)_assembly")[,3] 
)

#seqCo_sid is id assigned by sequencer company
#filename_sid is id assigned by LeBoa, a couple were skipped eg 42? 52?

# sanity check, looks good
# phylogroupResultDf |>   filter(str_detect(seq_filename, regex("42|52|72", ignore_case = TRUE)))

```

```{r QA_spot_explore_dat eval=F}

mlstResultTsv |> group_by( SCHEME ) |> summarise( n() )  # 1 schemes, mlst was ran with forced EC scheme.  # think already checked they were all E.coli (rMLST or something)

```

```{note eval=F}

# run rMLST on fasta that MLST said is not ecoli

```




```{r massage_input_file, eval=T}

metadataNfo |> str()
metadataNfo  |> select( sample_title ) |> unique()

metadataDF = metadataNfo |> 
  rename( sample_name = `*sample_name`, 
          strain      = `*strain`,
          User_Sample_Name = `User Sample Name`) |>
  mutate( sample_name_as_string = as.character(sample_name) ) |>   # manual conversion is better than forcing input type
  mutate(
    source_water   = if_else( grepl( "water",  sample_title, ignore.case=T),  1, 0, missing=0 ),
    source_produce = if_else( grepl( "lettuce|cilantro|strawberry|basil", sample_title, ignore.case=T),  1, 0, missing=0 ),
  ) |> mutate(
    QA_source_count = ( source_water + source_produce  )  # expect 1 for every row, check if not!
  )
  

metadata_hilite_col = c( "sample_name_as_string", "AI5395_Name", "User_Sample_Name",  "sample_title", "source_water", "source_produce", "QA_source_count" )

metadataDF |> filter( QA_source_count != 1 ) |> select( all_of( metadata_hilite_col ))  # they are indeed all 1, pattern match worked.
metadataDF |> select( all_of( metadata_hilite_col )) 
#metadataDF |> head(5)
#metadataDF |> View()
metadataDF |> str()


```




filenames eg: AI-1608_103_103_assembly.fasta
need to extract       ^^^
as join key to be able to merge metadata info.
use Regular Expression in str_match() to extract file id number


mlst output examples  (From Borbon eg)

mlst   AI-1608_206_206_assembly.fasta
AI-1608_206_206_assembly.fasta  ecoli   10      adk(10) fumC(11)        gyrB(4) icd(8)  mdh(8)  purA(8) recA(2)

mlst  --scheme ecoli AI-1608_206_206_assembly.fasta
AI-1608_206_206_assembly.fasta  ecoli   10      adk(10) fumC(11)        gyrB(4) icd(8)  mdh(8)  purA(8) recA(2)


mlst --legacy --scheme ecoli AI-1608_206_206_assembly.fasta
FILE    SCHEME  ST      adk     fumC    gyrB    icd     mdh     purA    recA
AI-1608_206_206_assembly.fasta  ecoli   10      10      11      4       8       8       8       2



--legacy is easiest to parse
grapetree and phyloviz likely just want the number (for matrix)

/mnt/c/tin/dataCache/Borbon_summer_25_Gabrielle/MLST_OUT_forceEC/mlst.all.forceEC.tsv


```{r prep_joinKeys}

mlstResultTsv |> str()

mlstResultDF_unfiltered = mlstResultTsv %>% mutate(
  seqCoSID = str_match( FILE, "AI[\\-]5395[\\_]([:digit:]+)[\\_]([:digit:]+)_assembly\\.fasta"  )[,2] ,  # sequencing company assigned sample id
  custSID = str_match( FILE, "AI[\\-]5395[\\_]([:digit:]+)[\\_]([:digit:]+)_assembly\\.fasta"  )[,3]     # customer sample id
  ) %>%  rename( forced_EC_scheme = SCHEME )   # to be very clear MLST output were all forced to E.coli

mlstResultDF_unfiltered |> filter( grepl( "4", custSID ) ) # customer sample id skipped = 42, 52, ...


mlstResultDF_unfiltered = mlstResultDF_unfiltered |> rename( isolateKey = custSID ) 

```


```{r join_metadata_and_mlst}

metadataDF |> str()
metadataDF |> dim()  # 74
metadataDF |> head()  # 
metadataDF |> select(  sample_name, sample_name_as_string, AI5395_Name, User_Sample_Name ) |> head()

mlstResultDF_unfiltered |> dim() # 72
mlstResultDF_unfiltered |> head()
mlstResultDF_unfiltered |> str()
mlstResultDF_unfiltered |> select( FILE, forced_EC_scheme, isolateKey ) |>  tail()  # 

meta_and_mlst_DF = full_join(
  x  = metadataDF,
  y  = mlstResultDF_unfiltered, #mlstResultDF,
  by = c( `sample_name_as_string` = "isolateKey" )
)


meta_and_mlst_DF %>% head()
meta_and_mlst_DF %>% tail()
meta_and_mlst_DF %>% dim()   # 74... check for NA in sample id++
#meta_and_mlst_DF |> View()   # still have House_ID info here   but Sample_Type_ID and WGS_ID , which as character, has .0 added.  weird!!

```





# massage VFDB table 
simplify, convert to gene present/absent
based on > 80% score as Match.

```{r getFirstNum, eval=T}

## abricate sometime have match value containing semicolon, eg
##  98.75;99.00
##  87.91;100.00;100.00;100.00;100.00
## 100.00;100.00;100.00;100.00             # usda_summary.csv:GCA_052631835.1_PDT002905087.1_genomic_usda.txt
## need to parse them out, maybe keep the first number, up to the first ;

## ref: Env_EcO157/code/stInfo_Snp1.Rmd
## ref: L348, L354.
## ./guatemala_amr/guat_amr_analy.Rmd:## semicolon shows up when multiple match in a given file.
## ./guatemala_amr/guat_amr_analy.Rmd:# many of the genes col could have semicolon separate numbers, eg


# abricate --sumary produce output where
# many of the genes col could have semicolon separate numbers, eg 98.75;99.00
# taking first ; ignore rest, convert to double.
# This kind of output happens when the file has multiple section where it matched.
# running the abricate --db jgrg db/amnt/sequence  will demo this output really well.
# define helper fn that will be used by across() inside a mutate().
getFirstNum = function( charStr ) {
  firstStr = str_match(  charStr , "([:digit:]+\\.*[:digit:]*)")[,1]
  if_else(
    as.numeric(  firstStr ) >= 80 , 1, 0, missing=0     # 80% as cutoff to Positive(1), else Negative(0)
  )
}
```




# join with abricate VFDB result.
# should do this earlier.

```{r massage_vfdb_result, eval=T}
#// vfdb_sum_df_BAK = vfdb_sum_df
#//usdaDB_sum_df = usdaDB_sum_df_text 

# VfdbResultTsv |> str() 

VfdbResultDf = VfdbResultTsv     |> 
  rename( file_name = `#FILE` )  |> mutate(
  seqCo_sid      = str_match( file_name, "AI-5395_([0-9]+)_([0-9]+)_assembly_Abricate_vfdb.tsv")[,2] ,   
  filename_sid   = str_match( file_name, "AI-5395_([0-9]+)_([0-9]+)_assembly_Abricate_vfdb.tsv")[,3]   ) |>
  mutate( across( 3:length( VfdbResultTsv), getFirstNum ))  

#  mutate( across( 4:length( VfdbResultTsv), getFirstNum ))  
#                 ^-- col idx and length()-1 are very fiddly.  depends on how read* fn worked?.  


#- VfdbResultDf_test = VfdbResultDf |> mutate( filename_sid   = str_match( file_name, "AI-1608_([0-9]+)_([0-9]+)_assembly_Abricate_vfdb.tsv")[,2]   )
#- VfdbResultDf_test |> View()
# extract sample ID from file_name, so that it can later be joined to meta data WGS_ID   eg 155
# file_name from vfdb result eg: 
#  AI-1608_155_155_assembly_Abricate_vfdb.tsv
  
VfdbResultTsv |> head(5)
VfdbResultDf  |> head(5)
#VfdbResultDf |> View()
VfdbResultTsv |> dim()    # 73x176
VfdbResultDf  |> dim()    # 73x178
VfdbResultDf  |> str()  
VfdbResultDf  |> select( length(VfdbResultDf)-170 : length(VfdbResultDf) ) |> str()
VfdbResultDf  |> select( 170 : length(VfdbResultDf) ) |> str()
VfdbResultDf  |> colnames()  

# Abricate result there is a funny recod 
# "AI-5393*_Abricate_vfdb.tsv 0 . . ."
# likely artifact from my run, or earlier run left over.  should not cause any damage "
```

# Check for presence of specific virulence genes, and define pathotype based on their presence/absence

++ double check gene names (in column names) eg is stx and not stx1 or stx1a , I think i checked column name when writting the check, which is why didn't use grepl, but it might have been for the previous project, double check for irr_exp26++  

See also https://docs.google.com/presentation/d/1boMR9VL7optd0HJf_z42qGZxgTt32oSNuTKzeiwYA9A/edit?slide=id.g3f6e9e62a90_0_34#slide=id.g3f6e9e62a90_0_34 slide 26 "Virulence of isolates" from LeBoa project update 


diffusely adherent E. coli (DAEC).	
entero Aggregative E. coli (EAEC)	
entero Invasive E. coli (EIEC)	
entero pathogenic E. coli (EPEC),	   #  aEPEC    atypical  vs  tEPEC"
entero toxigenic E. coli (ETEC),	
extraintestinal pathogenic E. coli (ExPEC)	##  uropathogenic E.coli UPEC, cause UTI, subset of ExPEC with most clinical signif.
Shiga toxin-producing E.coli	



file:///J:/My%20Drive/sn126_t6_phd_hw/Jay/UPEC%20local%20definition%20Quito%20Ecuador%202026.pdf


```{r determinePathotype}


pathotype_names = c( "ExPEC", "ETEC", "EPEC", "STEC", "EAEC", "DAEC", "EIEC" )  # "EIEC_Shigella",  "STEC_EHEC"

ETEC_genes = c( "lt", "sta" )
EPEC_genes = c( "eae" )         # this dataset has only eae
STEC_genes = c( "stx" )         # no stx* in vfdb result
EIEC_genes = c( "ipaH" )        # no ipa* in vfdb result
EAEC_genes = c( "aatA", "aggR", "aaiC" )        # OR condition, none exist in vfdb result
#DAEC_genes = c( "afa", "dra", "daa" )           # OR condition, names in VFDB result are more specific
DAEC_genes = c( "afaA", "afaB-I", "afaC-I", "afaD", "afaE-II", "afaF-III", "draA", "draB", "draC", "draD", "draE2", "draP", "daaD", "daaE", "daaF" )

# most gene definition from 
# doi:10.1097/QCO.0000000000000665   // my annotated file: file:///J:/Shared%20drives/Tin_Ho_gDrv_BL_shareDrv/mph_hw/mph_prj_bioinfo/ExPEC_Ala/Defining%20diarrheagenic%20E.%20coli.pdf 
# Table 1 p 16 
# DAEC said: Genes in the afa, dra, or daa adhesin operons; cross-reactivity with EAEC genes  (and that more work needed for better definition).  so just say predicted possible pathotype.
# Table 2 p 17, paper reported afaE-I or afaE-III


ExPEC_genes = c( "papA","papA","papB", "papC","papD","papE","papF","papG","papH" , "sfaC","sfaY", "kpsMII", "iutA" )   # only kpsM, no kpsMII

pathotypeDf = VfdbResultDf |>
  mutate(
    #ExPEC = if_else( papA + papC + sfaC + kpsMII + iutA >= 2, "MoreWorkNeeded", 0, missing=0 ),   # ExPEC definition: at least 2 of the 5 genes
    #ETEC  = if_else( lt + sta >= 1, "lt/ltA check", 0, missing=0 ),   # ETEC definition: at least 1 of the 2 genes   # no lt in this dataset
    #ETEC  = if_else(  if_any(any_of("lt"))  + if_any(any_of("sta")) >= 1, "lt/ltA check", 0, missing=0 ),   # ETEC definition: at least 1 of the 2 genes   # no lt in this dataset
    ExPEC = if_else(  rowSums( pick( any_of( ExPEC_genes)), na.rm=T  ) >= 2, 0.9, 0, missing=-1 ),     # ExPEC definition: at least 2 of 5 genes, and more nuanced... more work needed ++
    ETEC  = if_else(  rowSums( pick( any_of( ETEC_genes )), na.rm=T  ) >= 2, 1, 0, missing=-1 ),   # ETEC definition: at least 1 of the 2 genes   # this seems to work.  CHECK again: no lt or sta in (Burbon?) dataset.  ++
    EPEC  = if_else(  rowSums( pick( any_of( EPEC_genes )), na.rm=T  ) >= 1, 1, 0, missing=-1 ),
    STEC  = if_else(  rowSums( pick( any_of( STEC_genes )), na.rm=T  ) >= 1, 1, 0, missing=-1 ),
    EIEC  = if_else(  rowSums( pick( any_of( EIEC_genes )), na.rm=T  ) >= 1, 1, 0, missing=-1 ),
    EAEC  = if_else(  rowSums( pick( any_of( EAEC_genes )), na.rm=T  ) >= 1, 1, 0, missing=-1 ),   # or condition of 3 genes, thus still just checking for >=1 
    DAEC  = if_else(  rowSums( pick( any_of( DAEC_genes )), na.rm=T  ) >= 1, 1, 0, missing=-1 )    # or condition of 3 genes, thus still just checking for >=1 
    )

#pathotypeDf |> select( file_name, ETEC  )  |> head(10)
#pathotypeDf |> filter( ETEC != "No" ) |> select( file_name, ETEC  )  |> head(10)  # 0 rows
#pathotypeDf |> filter( EPEC != "No" ) |> select( file_name, EPEC  )  |> head(10)  # 1 rows

pathotypeDf |> filter( if_any( any_of( pathotype_names ), ~ .x != 0 )) |> select( file_name, filename_sid, any_of( pathotype_names)  )  |> head(25)  # 21 rows

#pathotypeDf |> View()

# while I was looking at the paper, copilot came up with the above.  i was going to use case when.  the gene names are wrong.  the ExPEC logic... did i use it before?
# search email ...
# https://doi.org/10.1038/s41579-020-0416-x  // dont look like local def
# Saba had a UPEC Local and Global... manuscript (not out yet) , it states: 
# study, a different molecular definition by Johnson et al. for the broader ExPEC group (presence
# of ≥2 of papAH and/or papC, sfa/focDE, afa/dra, iutA, kpsMII) was the strongest predictor of
# experimental virulence in a murine model.


# local def of ExPEC/UPEC: file:///J:/My%20Drive/sn126_t6_phd_hw/Jay/2026_Liseth_UPEC_manuscript_LS_jun26.pdf

# Saba's paper, UPEC local definition Quito Ecuador 2026   p2, details in page 15:
# file:///J:/My%20Drive/sn126_t6_phd_hw/Jay/UPEC%20local%20definition%20Quito%20Ecuador%202026.pdf
# Results: 
# The optimal local definition for ESBL-producing UPEC was the presence of ≥ 1 of upaH, cnf1, or sat

```


# save pathotypeDf   and color cell in spreadsheet to highlight the yes
# include Ceftriaxone status   so can just use spreadsheet to group/filter.  not too many records.



```{r pathotype_meta_join}

pathotype_meta_stage1 = full_join(
  x  = meta_and_mlst_DF, #meta_and_EC_ST,
  y  = pathotypeDf,  
  by = c( `sample_name_as_string`  = "filename_sid" )     # sample_name_as_string remains in resulting DF
  #by = c( "WGS_ID" = "filename_sid" )  
  #by = c( "FILE" = "file_name" )
) 

#pathotype_meta_stage1 |> str()
phylogroupResultDf |> str() 

# ++ join/add pathotypeResultTsv #0905
pathotype_meta_stage2 = full_join(
  x  = pathotype_meta_stage1, 
  y  = phylogroupResultDf,
  by = c(`sample_name_as_string` = "filename_sid")
)

# pathotype_meta_stage2 |> str()
# pathotype_meta_stage2 |> select( sample_name_as_string, phylogroup, file_name) |> print( n=5 )

pathotype_meta = pathotype_meta_stage2
# keeping previous DW_DF name so subsequent code don't need change #0905 phylogroup

#-- |> rename( CRO = CTX_class_binary )

  # |> mutate(
  # CRO = case_when( 
  #  grepl( "Ceftriaxone", Isolate_Type, ignore.case=T ) ~ "Positive",
  #  TRUE ~ "Negative" ) )
  # already have CTX_class_binary

# file_name from vfdb result eg: AI-1608_155_155_assembly_Abricate_vfdb.tsv
# FILE                eg AI-1608_103_103_assembly.fasta
# file_name           eg 
# fasta_used_by_rMLST eg AI-1608_103_103_assembly

# pathotype_meta |>  View()  # ~10 rows have NA, check with Gabrielle which NA are expected, which rows to filter, what to keep.



# sanity check

#--meta_and_EC_ST |> colnames()
#meta_and_mlst_DF |> colnames()

pathotype_meta |> str()    # __keyDF__

#pathotype_meta |> select( file_name, WGS_ID_as_string, any_of( pathotype_names) ) |> head(25)
#--pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, any_of( pathotype_names), CRO ) |> head(25)
#pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, forced_EC_scheme, `MLST results`, rMLST_specie, forced_EC_scheme, ST, any_of( pathotype_names), CRO ) |> View()
# pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, any_of( pathotype_names), CRO ) |> View()


#--pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, forced_EC_scheme, `MLST results`, rMLST_specie, forced_EC_scheme, ST, any_of( pathotype_names), CRO ) |> filter( is.na( House_ID))
# 
# 



```


```{r write_summary_pathotype_result, eval=F}

sum_pathotype_result_brief = pathotype_meta |> select( sample_name, sample_name_as_string, sample_title, any_of( pathotype_names) ) 
sum_pathotype_result_brief |> head()

#-- sum_pathotype_result_brief = pathotype_meta |> select( House_ID, WGS_ID_as_string, Isolate_Type, any_of( pathotype_names), CRO ) 

sum_pathotype_result = pathotype_meta |> select( sample_name, sample_name_as_string, seqCoSID, sample_title, forced_EC_scheme, forced_EC_scheme, ST, any_of( pathotype_names) ) 
sum_pathotype_result |> head()
# sum_pathotype_result |> View()

#~~ write_tsv( sum_pathotype_result, "../result/pathotype.tsv" )

# https://docs.google.com/spreadsheets/d/178qtzQ2wHiBp4ziJUgJUGHP9PUOf6FYw0fTXIvmy-50/edit?gid=749728428#gid=749728428 

#0820 draft result.  ExPEC need logic update.  rest of pathotype need review with Jay on logic being acceptable. ++



```

# https://docs.google.com/spreadsheets/d/178qtzQ2wHiBp4ziJUgJUGHP9PUOf6FYw0fTXIvmy-50/edit?gid=749728428#gid=749728428 

#0820 draft result.  ExPEC need logic update.  rest of pathotype need review with Jay on logic being acceptable. ++



have more columns from metadata and VFG that can be included as needed

Methods write up:


In-silico prediction of pathotype.  
Abricate is run on the WGS of the isolates to determine the virulence genes each isolate carries.
Genes that predict Pathotypes as described in Jesser and Levy (2020) [ doi:10.1097/QCO.0000000000000665 ] is screened for.
Briefly, 
presence of both _lt_ and _sta_ genes would be predictive of ETEC pathotype (none in this dataset).  
Presence of _eae_ genes would be predictive of EPEC pathotype (1 isolate in this dataset).
Presence of _stx_ genes would be preditive of STEC pathotype (none in this dataset).
Presence of _ipaH_ genes would be preditive of EIEC pathotype (none in this dataset).
Presence of any of _aatA_, _aggR_ or _aaiC_, genes would be preditive of EAEC pathotype (none in this dataset).
Presence of any of _afa_, _dra_ or _daa_ family of genes would be preditive of DAEC pathotype (~6 this dataset).

For ExPEC, we used a local definition (paper for reference) [More work TBD]

Please note that the pathotype is predicted from computational screening of DNA sequences, and, as stated by Jesser and Levy, some of these molecular based pathotype definition can use of further refinements.  Thus, please use these predicted pathotypes with caution.



(Figure caption, or even table headings, should emphasize it is predicted pathotype.)


```{r stopped_here_for_leboa}

# ^^ above should be good.  double check pathotype checked correctly would be nice.  ++
#    also don't have full ExPEC/UPEC def yet

# below is code originally for Gabrielle M Borbon data, not edited/adjusted for this project yet.  

```




```{r pathotype_prevalence, eval=T}
# Need a table as agreed, which have counts (and %)
# 

# also, check that 93 rows with data is the correct count ++


### make a smaller 2D_table first
#-- Patho_Source_DF = pathotype_meta |> select( House_ID, WGS_ID_as_string, source_mother, source_child, source_water, rMLST_specie, any_of( pathotype_names), CRO )
Patho_Source_DF = pathotype_meta |> select( sample_name, sample_name_as_string, seqCoSID, sample_title, source_water, source_produce, forced_EC_scheme, ST, any_of( pathotype_names ) )
#Patho_Source_DF |> View()

Patho_Source_DF_long = Patho_Source_DF |> pivot_longer( pathotype_names, names_to="Pathotype", values_to="Prevalence" )
# Patho_Source_DF_long |> View()

PathoPrevWater   = Patho_Source_DF_long |> filter( source_water    == 1 ) |> group_by( Pathotype ) |> summarize( Water_Prev   = sum(Prevalence))
PathoPrevProduce = Patho_Source_DF_long |> filter( source_produce  == 1 ) |> group_by( Pathotype ) |> summarize( Produce_Prev = sum(Prevalence))


PathoPrevProduce
# ^^ ++ Hmm... Produce_Pre doesn't seems right.  only NA?  ++FIXME++

PathoPrevTemp = left_join(
  x  = PathoPrevWater, 
  y  = PathoPrevProduce
)

PathoPrevAll = PathoPrevTemp
  # do additional join if more categories than water, produce

PathoPrevAll    #0820   this table decent, pasted to gSheet for discussion
# 


# ++ OH... Need to check denominator count.  E.coli?  what rows had been filtered?  ++

# if desire to find prevalence count by CRO +ve vs -ve, 
# do a Patho_Source_DF_long |> filter( CRO ... ) 
# and recreate the table.

```



                    ++CHECK++

# rest still Gabrielle's Brobon code.  it was abandoned.  unlikely needed here.



```{r double_pivot_longer_create_problem__dont_use, eval=F}
Patho_Source_DF_long_src = Patho_Source_DF_long |> pivot_longer( c(`source_mother`, `source_child`, `source_water` ), names_to="Source", values_to="Source_count" ) # dont like name...
#^^ this second pivot may be messing with values.  creating extra counts... thus the problem.
#^^ instead of creating 3 cols for source_mother, source_child, source_water, create a single col with 3 names, then don't have to pivot twice.  #0614



Patho_Source_DF_long_src  |> group_by( Pathotype, Source ) |> summarize( PathoPrev = sum(Prevalence))
#^^ result is wrong this way!!   sum() didn't split by the 2nd group_by... 

# deleting rest of pathotype prevalence count code that didn't produce correct result, 
# cuz the 2nd pivot_long induced extra counts and corrupt the process
# if needed, see git log e7b7ca2  for the old code
```

========================================
========================================
========================================
~~~~ Part 2 - tree (not much about pathotype, other that it would use that info)
========================================
========================================
========================================

# create a df that contains all the descriptive data for each isolate

df suitable to generate heatmap, which can be added to a ggtree



```{r QA_check_data_before_creating_heatmap_df, eval=F}
# dont need to run this after get code debugged

meta_and_mlst_DF |> colnames()
# pathotypeDf      |> colnames()
# pathotype_meta |> colnames()

meta_and_mlst_DF |> filter( grepl( "42|52", seqFileName  ))
pathotype_meta   |> filter( grepl( "42|52", seqFileName  ))
pathotype_meta |> filter(if_any(everything(), is.na))
# the DAEC, ETEC stuff need massage into single column containing the various type before it can be joined to the heatmap-DF

meta_and_mlst_DF |> select(ST) |> unique() |> count()           # 43 ST, color in heattile a bit hard to read
meta_and_mlst_DF |> group_by( ST ) |> summarize( freq = n() )
```

```{r df4heatmap_descriptive_data}


df4heatmap_descriptive_data_all = pathotype_meta |> 
  drop_na() |>
  mutate( label_id = stringr::str_c( "EC", sample_name_as_string,  sep="_")) |>  # cuz I named the gff EC_<SampleID> and those became the tree tip lable, so need to create this column to join things to tree
  select( label_id, ST, sample_title   ) 
  # sample_title is source, eg lettuce, water
  # select( sample_name_as_string, ST   ) 

#^^ add more col later, but may need to read more input file first ++

## ref // done at higher up // pathotype_meta = full_join(   x  = meta_and_mlst_DF,   y  = pathotypeDf,    by = c( `sample_name_as_string`  = "filename_sid" )  )

# df4heatmap_descriptive_data |> str()  # 72 rows after drop_na(), 75 before.  so 3 rows dropped, as desired.

# df4heatmap_descriptive_data  # wanted an _all and _some_subset version?


df4heatmap_descriptive_data_source = df4heatmap_descriptive_data_all |> 
  select( label_id, sample_title   ) 

```

# phylotree

```{r tree}

#-- treeFile = "../result/RAxML_bestTree.ECirr_raxml.tre"
#^^ tmp here, defined near top

p_load(phytools)

tree = read.tree( treeFile )



```


```{r some_very_simple_tree_sanity_check_mostly}
# The simplest class of visualization method simply plots the data at the tips of the tree. For instance:
# dotTree(tree, NULL,length=10,ftype="i")

p_load(ggtree)

#xx dotTree(tree_cg, geo_cgST_year_heat_mtx, length=10, ftype="i")

#slide103 = 
  ggtree(tree, layout = "fan") + geom_tiplab(aes(label = label, align=T))
         # fan = circular layout; align do nothing in circular layout.

    ggtree(tree, ) + geom_tiplab(aes(label = label, align=T))

  
```

```{r ggtree_plus_dat}
# ref: https://www.epirhandbook.com/en/new_pages/phylogenetic_trees.html#simple-tree-plus-sample-data


# Prepare data for gheatmap (rownames must match tip labels, i.e. the label column)
heatmap_df = as.data.frame(df4heatmap_descriptive_data_all)


rownames(heatmap_df) = heatmap_df$label
#~~old ref~~ heatmap_df_clean = heatmap_df %>%   select(-cgST, -label, -label2, -label2_info)
heatmap_df_clean = heatmap_df |> select( -label_id )

# rownames(heatmap_df_clean)  # just checking

tree # maybe need to fix tip label to just be the sample id number, end up creating df for heatmap with EC_ prefix to match tree.

tree_phylo = as.phylo( tree )
tree_phylo   # 72 tip, so that's right, should not have a Sample ID 42 , 52


#?? tree_annot = left_join(  x  = tree_phylo,    y  = heatmap_df_clean,   by = "label"  )

## things aren't joined at the right order / label / sample id ++FIXME++

# Base tree plot - display label (the cg-number) as node/tip label
p_tree = ggtree(tree_phylo) +
  geom_tiplab(aes(label = label), size = 2.5, align = TRUE, linetype = "dotted", linesize = 0.1)

# Add the heatmap layer of counts next to the tree tips
fig_gtre_1 = gheatmap(p_tree, heatmap_df_clean )
                     # offset =  0.03 )  

fig_gtre_1


```
```{r DBG_L774, eval=F}

# tmp debug code
#df4heatmap_descriptive_data |> dim()
#df4heatmap_descriptive_data |> filter(if_any(everything(), is.na))
#df4heatmap_descriptive_data |> filter( ST == "-" )
#df4heatmap_descriptive_data |> filter( is.na(ST) )    #  sample 42, 52 and a NA should no longer be here after drop_na() above


#df4heatmap_descriptive_data |> class()
#heatmap_df |> class()
#heatmap_df |> str()
 
#heatmap_df$label |> unique() # 81 uniq

```
```{r alt_display_tree}

p_tree_fan = ggtree(tree_phylo, layout="fan") +
  geom_tiplab(aes(label = label), size = 2.5, align = TRUE, linetype = "dotted", linesize = 0.1) 

fig_gtre_1a = gheatmap(p_tree_fan, heatmap_df_clean )
fig_gtre_1a

```


too many ST, color coding for heatmap is confusing, dropping it as suggestions.

```{r tree_no_ST}

# so source only

heatmap_df_noST = as.data.frame(df4heatmap_descriptive_data_source)

rownames(heatmap_df_noST) = heatmap_df_noST$label
#~~old ref~~ heatmap_df_clean = heatmap_df %>%   select(-cgST, -label, -label2, -label2_info)
heatmap_df_noST_clean = heatmap_df_noST |> select( -label_id )

# rownames(heatmap_df_noST_clean)  # just checking

fig_gtre_2a = gheatmap(p_tree_fan, heatmap_df_noST_clean )
fig_gtre_2a


```


```{r QA_spot_check}

# pick some id or ST and verify diagram is done right

meta_and_mlst_DF |> select(ST) |> unique() |> count()           # 43 ST, color in heattile a bit hard to read
meta_and_mlst_DF |> group_by( ST ) |> summarize( freq = n() )

meta_and_mlst_DF |> filter( ST=="10391" )
meta_and_mlst_DF |> filter( sample_title=="strawberry" )
meta_and_mlst_DF |> filter( sample_title=="basil" )
meta_and_mlst_DF |> filter( sample_title=="cilantro" )  # source match EC_<SampleID> in plot
meta_and_mlst_DF |> filter( sample_title=="lettuce" )

meta_and_mlst_DF |> filter( sample_title=="farm irrigation water" )
```

```{r date}
# R rated date :D
#Sys.Date()
Sys.time()

# bash date doesn't show in the R console, so this is better marker.
```

```{bash litmus_test}

date

```


